Interplay between homeostatic synaptic scaling and homeostatic structural plasticity maintains the robust firing rate of neural networks

Version 1

Abstract

Abstract

Critical network states and neural plasticity are essential for flexible behavior in an ever-changing environment, which allows for efficient information processing and experience-based learning. Synaptic-weight-based Hebbian plasticity and homeostatic synaptic scaling were considered the key players in enabling memory while stabilizing network dynamics. However, spine-number-based structural plasticity is not consistently reported as a homeostatic mechanism, leading to an insufficient under-standing of its functional impact. Here, we combined live-cell microscopy of eGPF-tagged neurons in organotypic entorhinal-hippocampal tissue cultures and computational modeling to study the re-sponse of structural plasticity under activity perturbations and its interplay with homeostatic synaptic scaling. By following individual dendritic segments, we demonstrated that the inhibition of excitatory neurotransmission did not linearly regulate dendritic spine density: Inhibition of AMPA receptors with a low concentration of 2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline (NBQX, 200 nM) sig-nificantly increased the spine density while complete blockade of AMPA receptors with 50 µM NBQX reduced spine density. Motivated by these results, we established network simulations in which a biphasic structural plasticity rule governs the activity-dependent formation of synapses. We showed that this bi-phasic rule maintained neural activity homeostasis upon stimulation and permitted both synapse formation and synapse loss, depending on the degree of activity deprivation. Homeostatic synaptic scaling affected the recurrent connectivity, modulated the network activity, and influenced the outcome of structural plasticity. It reduced stimulation-triggered homeostatic synapse loss by downscaling synaptic weights; meanwhile, it rescued silencing-induced synapse degeneration by am-plifying recurrent inputs via upscaling to reactivate silent neurons. Their interplay explains divergent results obtained in varied experimental settings. In summary, calcium-based synaptic scaling and homeostatic structural plasticity rules compete and compensate one another other to achieve an eco-nomical and robust control of firing rate homeostasis.

Significance Statement

This work combined in vitro experiments and computer simulations to study the interplay between homeostatic structural plasticity and synaptic scaling.

We observed a non-linear relationship between spine numbers and neural activity, where partial or complete inhibition of synaptic transmissions led to increased and reduced spine density, respec-tively.

Partial inhibition increased spine sizes regardless of their initial sizes, while complete inhibition selectively increased relatively large spines.

A bi-phasic spine-number-based structural plasticity rule reconciled the divergent experimental results in activity-dependent spine density changes.

We used a framework of engineering systems and complex systems to analyze the roles of the bi-phasic structural plasticity in maintaining the robustness of the neural network — firing rate homeostasis.

The bi-phasic rule partially uses the negative feedback strategy and serves as a redundant and heterogeneous mechanism to the synaptic-weight-based homeostatic synaptic scaling rule.

Both rules are based on intracellular calcium concentration which is the integral signal of neural activity. We, therefore, proposed a critical role for integral feedback control in firing rate home-ostasis.

Introduction

To survive in a dynamic world, animals must properly respond to novel and familiar environmental cues. The former demands their brain networks to sustain activity at a critical state, where information can be transmitted into action potentials and conveyed as neural avalanches.1–6 The latter describes experience-based learning such that subtle memory cues could trigger excessive responses in the corresponding pathway during recall. These two mechanisms are reflected by firing rate homeostasis and associative learning, respectively.

Robust firing rate homeostasis has been reported in rodents’ visual cortices and hippocampi in vivo, where neural activities get restored at either individual neuron7–9 or population level10, 11 within a few days after perturbation. Long before that, however, the idea of firing rate homeostasis has been hover-ing in theories since long-term synaptic potentiation (LTP) was observed.12 LTP is a positive-feedback modification of synaptic transmission among excitatory neurons (“neurons that fire together, wire to-gether”13), which was postulated by Donald Hebb14 and thus termed Hebbian plasticity. Its associative properties allow synapses to preserve memory traces as increased synaptic weights but risk pulling the network dynamics from equilibrium status into overexcitation or silence.15 Restoring the neural network to its equilibrium dynamics without erasing memories in this condition is therefore critical and attests to the robustness of the complex system—the human brain network.16–21

Robustness is a ubiquitous concept in many complex systems or engineering systems, meaning that when external perturbations forcefully change its dynamics, the system adapts to return to its original attractor or moves to a new attractor state that maintains its functions.19, 20 Several features have been proven essential for maintaining robustness, including negative feedback control, redundancy, and heterogeneity. Homeostatic synaptic scaling falls perfectly into the negative feedback strategy.22 Upon long-term manipulation of neural activity,22 membrane potential,23 or calcium concentration,24 neurons proportionally down-or upscale their incoming synaptic weights in a compensatory way. Redundancy and heterogeneity encompass various mechanisms that achieve the same goal in case one or two fail. While redundancy and heterogeneity have been identified in biological systems in the form of genetic buffering and convergent molecular circuits,19, 20, 25 their representation in activity-dependent plasticity has not been discussed explicitly. Induction of homosynaptic long-term depression (LTD)26 with the Bienenstock– Cooper–Munro (BCM) rule27 or STDP rule,28–31 or induction of heterosynaptic LTD via synaptic tagging and capture32 are in agreement with the idea of stabilizing firing rate. Inhibitory plasticity also contributes to re-establishing a balance between excitation and inhibition.33–35 All these mechanisms support the notion that in the functional plasticity regime, there are local or global mechanisms that are redundant and heterogeneous to the homeostatic synaptic scaling.

However, whether such redundancy is also reflected in the structural substrate of functional trans-mission remains elusive. Structural plasticity includes changes in sizes or numbers of dendritic spines and axonal boutons, numbers of synapses, and network connectivity, which are all pivotal to functional transmission and memory capacity.36 Meanwhile, previous work on activity-dependent changes in spine density showed both homeostatic and non-homeostatic regulations upon activity perturbation.37 In the present study, we used whole-cell patch-clamp recordings and time-lapse imaging experiments, which in-formed the computational implementation of a bi-phasic activity-dependent (i.e., calcium-based) rule for synapse-number-based structural plasticity that reconciles various alterations of spine numbers. Then, with computer simulations, we showed that this structural plasticity rule is competitive, redundant, and compensatory to the synaptic-weight-based homeostatic synaptic scaling rule, in different time scales and magnitude scenarios. We demonstrated that calcium concentration-based integral feedback control, in the formats of structural plasticity and synaptic scaling, enables robust adaption of network activity.21, 38, 39

Results

Integral feedback mechanisms in regulating firing rate homeostasis

Neural network dynamics are determined by the product of external inputs and intrinsic connectiv-ity. External stimulation and inhibition increase or decrease neural activity deterministically in a static network. Hebbian plasticity amplifies network activity by potentiating synaptic weights via positive feedback. In contrast, homeostatic synaptic scaling and homeostatic structural plasticity reduce input-triggered responses by modifying synaptic weights or numbers in the presence or absence of Hebbian plasticity (Fig. 1A). Mechanism-wise, homeostatic synaptic scaling and homeostatic structural plasticity seem redundant in regulating network dynamics. However, compared to homeostatic synaptic scaling, ex-perimental spine density analyses yielded ambiguous, even contradicting results, especially upon chronic activity inhibition37 (Fig. 1B). Highly heterogeneous experimental systems partially explain the inconclu-sive results. Nevertheless, in a complex biological neural network where homeostatic synaptic scaling is constantly subject to and meanwhile modulates neural activity, changes in synaptic weights are expected to interact with the activity-dependent structural remodeling of networks. From the perspective of a com-plex system, we proposed to view synaptic scaling and spine-number-based structural plasticity as two integral feedback mechanisms (Fig. 1C), which use intracellular calcium concentration (C(t)) to inform their regulatory behaviors. We approximated their interactions systematically using computer simula-tions. Specifically for structural plasticity, different working models are available where neural activity either monotonically or non-monotonically governs the neurites’ outgrowth and retraction, and hence synapse formation and loss. Additional information is therefore needed to inform the model selection.

Integral feedback mechanisms in regulating firing rate homeostasis.

(Ai) Neural net-work activity is determined by external inputs in a static network. (Aii) Hebbian plasticity amplifies the network responses to external inputs via a positive feedback mechanism. (Aiii) Homeostatic plasticity restores setpoint activity via a negative feedback mechanism, which is key to firing rate homeostasis. (Bi) Hebbian functional plasticity strengthens recurrent connectivity by potentiating the weights of certain groups of synapses. (Bii) Homeostatic synaptic scaling proportionally down-scale or up-scale all synaptic strengths upon chronic excitation and inhibition. (Biii) Homeostatic structural plasticity presents homeo-static spine loss upon chronic excitation, while divergent changes in spine density upon chronic inhibition have been observed. (C) Synaptic scaling and structural plasticity use intracellular calcium concentration ([Ca2+]i, C(t)) to track neural activity (AP, action potential, S(t)). Calcium concentration updates each time with calcium influx (βCa) upon the arrival of an action potential and decays with a time constant τCa. Homeostatic synaptic scaling is implemented as a weight-dependent rule which updates the synaptic weight w(t) with a scaling factor ρ. The discrepancy from the setpoint ɛ determines the direction of weight scaling. Structural plasticity is also calcium-dependent and governs the growth and retraction of axonal boutons and dendritic spines by the setpoint value. Two examples of the structural plasticity rule are presented here. They are either linearly or non-linearly dependent on the intracellular calcium concentration.

Heterogeneous modification of spine sizes and numbers upon chronic synaptic inhibition

To examine whether spine density is linearly dependent on neural activity or not, we used 200 nM and 50 µM NBQX (2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline), a competitive AMPA receptor antago-nist, and studied CA1 pyramidal neurons of entorhinal-hippocampal tissue cultures (Fig. 2A). Whole-cell patch-clamp recording of AMPA receptor-mediated spontaneous excitatory postsynaptic currents (sEP-SCs) demonstrated a significant reduction in mean sEPSC amplitude and frequency with 200 nM NBQX and a near-to-complete blockade of excitatory synaptic transmission with 50 µM NBQX (Fig. 2Bi-Bii). Then the two concentrations were applied to Thy1-eGFP cultures for three days, and individual dendritic segments were followed before and after chronic activity inhibition with the help of time-lapse imaging (Fig. 2C). Analysis showed that 200 nM NBQX treatment increased spine density (p = 0.003, Wilcoxon test), while 50 µM NBQX treatment reduced spine density (p = 0.008, Wilcoxon test), in comparison to their baseline densities (Fig. 2D). No significant changes were observed in the control cultures (p = 0.06, Wilcoxon test). Our results indicated a non-linear activity dependency for spine numbers.

Heterogeneous modification of spine numbers and sizes upon three-day synaptic inhibition via NBQX.

(A) Example CA1 pyramidal neuron recorded in the entorhinal-hippocampal tissue culture for probing the effects of different NBQX concentrations. Scale bar: 500 µm. (Bi-Bii) Frequency and averaged amplitudes of sEPSC events for each recorded neuron in three groups (N = 18 for the control group, N = 18 for the 200 nM NBQX-treated group, N = 12 for the 50 µM NBQX-treated group). (C) Example Thy1-eGPF culture and the example dendritic segments from the radiatum layer (rad.) before and after three-day treatment. Scale bar: 200 µm and 5 µm. (D) Spine density at baseline and after the three-day treatment. All values were normalized by the corresponding baseline values. Lines with light shades are raw data (solid and dashed lines represent increased or reduced spine density, respectively). Dark-shaded lines with error bars are each group’s means and the standard error of the means (s.e.m.s). (N = 19 for the control group, N = 24 for the 200 nM-treated group, N = 33 for the 50 µM-treated group) (E) Cumulative distribution function of spine sizes before and after the three-day treatment (N = 489 for the control group, N = 736 for the 200 nM-treated group, N = 675 for 50 µM-treated group). Inset shows the corresponding averages of normalized spine sizes. (F) Normalized changes in spine sizes grouped by their initial spine sizes. Values on the x-axis are the upper limits of each group. (G) Each segment’s average change in spine sizes against its initial spine density. The marker size labels the net change in spine density over a three-day course. (H) Each segment’s average changes in spine sizes against its change in spine density. (I) The table summarises the experimental data, while the graph displays the extrapolated relationship between neural activity and spine densities or sizes.

We also tracked the size of individual spines before and after the treatment. The spine sizes in all three groups followed a long tail distribution in which the majority were small (Supplementary Fig. 1A) while their alterations after the three-day treatment (Δ spine size) displayed a normal distribution (Supplemen-tary Fig. 1B). Population-wise (Fig. 2E), we observed a significant increase and a significant reduction in spine sizes, respectively, after 200 nM (p < 0.001, Wilcoxon test; 99.9%CI = [9.09, 25.8], LLM) and 50 µM NBQX treatment (p < 0.001, Wilcoxon test; 99.9%CI = [9.09, 25.8], LLM). However, we noticed that after normalizing the size alterations by their initial sizes, all three groups showed an increased ten-dency (Fig. 2E inset and Supplementary Fig. 1C), which we suspected was inflated by the enlargement of large numbers of small spines. Therefore, we sorted the spine size alterations based on their initial sizes in Fig. 2F. Indeed, control segments (black curve) showed a natural fluctuation of spine sizes that small spines tend to grow over the three-day period while large spines shrink. When partial inhibition via 200 nM NBQX (orange curve) was applied, the overall dynamics shifted toward increase regardless of their initial sizes. However, complete inhibition via 50 µM NBQX (light blue curve) only tended to increase the sizes of a small number of large spines, but counterbalanced the natural increase of small spines and caused shrinkage of middle-sized spines. These data confirmed a general homeostatic adjust-ment of spine sizes upon partial and complete inhibition. However, their initial sizes seem to mediate the fate of individual spines in response to complete inhibition.

To provide insights into the interaction between spine density and size, which may hint at the rela-tionship between structural plasticity and synaptic scaling, we further quantified the average changes of all spine sizes for individual dendritic segments (spine size). We observed a positive correlation be-tween the average changes in spine sizes and their baseline spine densities under 200 nM NBQX treatment (Fig. 2G, r(24) = 0.65, p = 0.00064, Pearson’s correlation; Supplementary Fig. 2A) but not in control or 50 µM NBQX treated groups. When plotting the average changes of spine sizes against the corre-sponding alterations of spine densities for all dendritic segments, we found that weak inhibition (200 nM NBQX) triggered both spine density increase and spine size enlargement in most of the segments (see orange dots clustered in the upper right quadrant in Fig. 2H). These data implied a potentially critical role for segments with dense spines and a synergistic effect between structural and functional plasticity upon weak inhibition. In the case of 50 µM NBQX treatment, however, the size-density data points were widely scattered in four quadrants (light blue dots in Fig. 2H). This result supported our observation of heterogeneous modification of spine sizes upon strong inhibition. In summary, our imaging results suggested a non-linear relationship between network activity and spine density or spine sizes, which may result from the complex interaction between synaptic scaling and structural plasticity upon weak and strong activity deprivation (Fig. 2I).

Stabilizing and characterizing a biphasic structural plasticity rule

Our results regarding spine density alterations demonstrated a non-linear (biphasic) relationship between neural activity and spine numbers, where partial inhibition resulted in an increase in spine numbers and complete inhibition resulted in reduced spine numbers. Because changes in spine numbers or density are functionally correlated to synapse formation, loss, and rewiring, we decided to implement structural plasticity rules in point neurons to study synaptic rewiring in a large homogeneous spiking neuron network without considering the heterogeneous neural morphology (Fig. 3A). The vanilla linear growth rule was implemented for comparison reasons (Fig. 3Bi). We adopted a Gaussian-shaped growth rule of synaptic element numbers with two setpoints to represent the observed biphasic dependency for simplification. Particularly, by setting the first setpoint at η = 0 or η > 0, we achieved two slightly different variants of the Gaussian rule (Fig. 3Bii-Biii). We then grew three neural networks where the synapses among excitatory neurons were respectively subject to the three mentioned rules (Fig. 1C). As reported before,11, 40–42 the linear-rule-guided network developed smoothly to a homogeneous sparsely connected network (10% of connection probability, equivalent to 1 000 excitatory synapses for each neuron) with an average firing rate around the target rate, i.e., the setpoint value ɛ = 7.9 Hz (Fig. 1Di). The Gaussian rule with a zero setpoint (η = 0) also developed the network to an equilibrium state similar to the linear rule (Fig. 1Dii). However, the Gaussian rule with two non-zero setpoints (η = 0.7 and ɛ = 7.9) did not permit proper network development (Fig. 1Diii).

Growing a neural network with three distinct structural plasticity rules.

(A) The point neuron model was used to study structural plasticity, where dendritic morphology is reduced. Dendritic spines are represented by pink sticks on the soma; axonal boutons are represented by empty or solid half circles. An empty circle with a dashed line labels an axon during retraction. Calcium concentration is linearly correlated with neural firing rate in our implementation, so neural activity would be used in the rest of the manuscript to reflect the hidden calcium dynamics. (Bi-Biii) Three activity-dependent growth rules of structural plasticity regulate the change of synaptic element numbers. (Bi) Linear rule with one setpoint (ɛ = 7.9). (Bii) Gaussian rule with two setpoints that one is zero (ɛ = 7.9 and η = 0). (Biii) Gaussian rule with two non-zero setpoints (ɛ = 7.9 and η = 0.7). Three shades indicate 100%, 50%, or 10% of the original growth rate (ν). Positive and negative values indicate, respectively, the speed of outgrowth and retraction of synaptic elements. (C) The neural network architecture of the Brunel network. 10 000 excitatory (blue) and 2 500 inhibitory neurons (red) are stimulated by external Poissonian inputs. All I-I, I-E, and E-I synapses are hard-wired with 10% probability. E-E synapses are subject to structural plasticity rules. (Di-Diii) Temporal dynamics of neural activity and network connectivity (Γ) during growth, respectively, guided by three rules. If not otherwise stated, the curve and the shaded area in activity plots represent the mean and standard deviation of the neural activity for the inhibitory population (I) and excitatory population (E). The network developed to an equilibrium state (Γ = 10%) in Di and Dii but not in Diii. The firing rates distribution and network connectivity matrices of the chosen time points, indicated by solid triangles, are included in Supplementary Figure 3 for Di-Dii, and in Figure 4 for Diii.

A closer scrutinization of the neural firing rates and network connectivity at a late growth stage showed that half of the excitatory neurons were silent (Fig. 4A), and all excitatory neurons were connected inhomogeneously where almost half of them were isolated (Fig. 4Bi-Bii). Correlating the firing rates and synapse numbers of individual excitatory neurons confirmed that the silent neurons were isolated while neurons that reached the target rate possessed a regular synapse number close to that observed in the linear rule (1 000 synapses, Fig. 4C). These observations are consistent with the properties of the Gaussian growth rules, indicating that one setpoint is stable (ɛ = 7.9), while the other one is either stable if the synapse numbers remain unchanged after silencing (η = 0) or unstable if disconnection occurs after silencing (η > 0). We thus refer to the Gaussian rule with two stable setpoints as stable Gaussian rule and the other Gaussian as biphasic Gaussian rule in the manuscript.

Silent neurons remained isolated in the network regulated by the biphasic Gaussian rule.

(A) Histogram of firing rates of excitatory and inhibitory neurons sampled at the time point indicated in Figure 3Diii. Almost half excitatory population was silent. The blue vertical line labels the mean firing rate of non-silent neurons. The orange dashed vertical line indicates the target firing rate (ɛ = 7.9 Hz). (Bi-Bii) Network connectivity matrix and the distribution of synapse numbers that individual excitatory neurons have. (C) Correlation heatmap between neural activity and synapse number of individual excitatory neurons. Neurons that were silent did not form synapses either. Neurons that fired around the target rate formed around 1 000 synapses from other active excitatory neurons. (D) Network architecture when facilitating current (Ifacilitating) was injected to boost the network development. (E) Temporal dynamics of neural activity and network connectivity when damping facilitating current were injected. The small inset shows the firing rate distributions of both excitatory and inhibitory neurons at a chosen time point (solid triangle). The facilitating current started at 750 pA and decayed linearly to zero at 4 000 s. (F) Different starting values of facilitating currents ended with different network connectivities. We used 750 pA throughout the manuscript.

Inspired by the observation that neural excitability facilitates circuit development,43 we increased the excitability of excitatory neurons during the growth period by injecting facilitating current (Ifacilitating) to lift the average membrane potential closer to the threshold potential (Fig. 4D). With the help of facilitating current, the neural network with biphasic Gaussian rule grew to the equilibrium state and retained the dynamics after the facilitating current decayed to zero (Fig. 4E). Given that different inten-sities of facilitating current resulted in different equilibrium connectivity (Fig. 4F), we used the intensity that achieved the same network connectivity (10%) and firing dynamics as the linear rule for further exploration.

Biphasic Gaussian rule reconciles divergent structural changes triggered by activity perturbation

As a proof-of-concept experiment, we systematically tuned the input strength to a subgroup of excitatory neurons (10% of the total excitatory population) from 0% to 200% fold of the original intensity (FOI, Fig. 5A) after the growth period to mimic input deprivation and stimulation. Example traces in Fig. 5Bi-Biii,C showed the protocols, neural activity, and network connectivity when the network was subject to the biphasic Gaussian rule. Intuitively, stimulation increased the neural activity of the subpopulation (S) (left panel in Bii) and triggered a reduction in synaptic connectivity among the stimulated neurons (S-S) and between the stimulated and non-stimulated excitatory neurons (S-E; left panel in Biii). The connectivity matrices in panel C display the final network connectivity. The homeostatic disconnection eventually restored the activity level of the stimulated neurons to the setpoint value (orange line in Bii). In contrast, the magnitude of activity deprivation determined whether the connectivity underwent homeostatic increase or silencing-induced degeneration. Both weak deprivation and silencing decreased the neural activity of the subpopulation (middle and right panels in Bii), while the network only managed to restore its network activity to the setpoint value by homeostatically increasing the network connectivity under weak deprivation. Silencing protocol induced disconnection of the silent neurons (right panels in Biii and C).

Divergent regulation of network connectivity upon stimulation and deprivation by three structural plasticity rules.

(A) A subpopulation (10%) of excitatory neurons (S) was subject to activity perturbation. All E-E, S-S, and E-S synapses are subject to the biphasic Gaussian rule. (Bi) Activity perturbation protocol. Three different folds of the original intensity (FOI) of the Poisson generator were used as examples to represent stimulation (110% FOI), weak deprivation (95% FOI), and silencing (0% FOI). (Bii) Temporal dynamics of neural activity of the subpopulation (S), the rest of the excitatory neurons (E), and inhibitory neurons (I) under corresponding protocols. (Biii) Temporal evolution of the overall network connectivity and connectivity of different subgroups under corresponding protocols. Synaptic connection probability from E to S (E-S) is identical to that from S to E (S-E) here so only S-E traces are shown. (C) Network connectivity matrices at the end of three protocols. (D) Average incoming synapse numbers of S neurons under different FOIs. Empty green circles are data from networks under extreme stimulation or inhibition, where neural activity and network connectivity dynamics were unstable. (E) Examples of two neurons that received the same external inputs but have different calcium decay time constants (τCa). The upper panel shows the membrane potential; the middle panel shows the spike trains of the two neurons; the lower panel displays the integrated calcium concentration over time. (F) Connectivity traces of subnetwork upon silencing under three different conditions.

Systematic analysis of incoming synapse numbers confirmed the biphasic dependency of the biphasic Gaussian rule (dark yellow curve in Fig. 5D) and the monotonic property of the linear rule (green curve). The stable Gaussian rule, which has two stable setpoints (η = 0), however, showed an intermediate tendency that external stimulation and weak deprivation triggered a homeostatic reduction or increase in synapse numbers, whereas strong deprivation left the network silent and intact (light yellow curve). Among the three rules, the biphasic Gaussian captured the homeostatic properties of the linear rule and allowed for silencing-induced spine loss, as reported in our experiments and many other previous experiments.44–48

Lesion-or denervation-induced plasticity is of considerable interest in activity-dependent structural plasticity and clinical-relevant scenarios.49–51 The biphasic Gaussian rule thus provides a model to study denervation-induced degeneration and subsequent regeneration. Nevertheless, in the temporal evolution of network connectivity, we noticed a small increment right after the silencing, which seemed biologically unrealistic. This might have resulted from the fast growth rate or the residual effects of previous firing activity on calcium concentration. In Fig. 5E, we showed that two neurons receiving the same external inputs presented identical membrane potential dynamics and spike trains, while the difference in calcium time constant (τCa) led to a gap in accumulated calcium concentration. Because the synaptic element number depends on the calcium concentration, which reflects the firing rate, the shorter time constant (τCa = 1 s, orange curve in 5F) carried less activity history and resulted in a smoother reduction of synaptic connectivity. Similar effects were achieved by using a slower growth rate (10% of the original rate) for structural plasticity (light blue dashed curve, 5F). Here we showed with computer simulations that the time scales of calcium dynamics and structural plasticity interfered. In reality, the decay time constant of calcium concentration is difficult to estimate due to multiple sources of calcium signals across time and space, while structural plasticity is widely accepted as a rather slow process occurring within hours to days. Therefore, we decided to use the slow growth rate and the original calcium time constant to further study the activity-dependent structural plasticity. The fast growth rate was only used in Fig. 6D-F to accelerate spine turnover in order to save simulation time when we need to display the dynamics over a longer period.

Activity perturbation and recurrent connectivity shaped the evolution of network connectivity.

(A) In the default network, we used the same growth rule for axonal boutons (A, light brown curve) and dendritic spines (D, pink curve). Alternatively, different η values could be used for axonal and dendritic elements. A silencing protocol was applied. (B) Neural activity of S, E, and I neurons under two conditions. (C) The network connectivity matrices at the end of the silencing protocol under two conditions. (D) Protocols used to examine the effects of recurrent connectivity and external stimulation. (E-F) Time courses of neural activity and connectivity upon silencing and external stimulation. Under Protocol 1, the growth rate in the left panel was ten times faster than that in the middle panel. In the right panel, external stimulation intensity was doubled.

Activity perturbation and recurrent connectivity shape the evolution of spine number

Although silencing-induced degeneration captured the feature of deprivation-induced spine loss, it was rarely observed that input deprivation led to complete neural isolation. Rather, spine recovery has been observed after spine loss.44, 46, 48 Early computer simulations with the Gaussian rule also reported such “physiological” recovery phenomena by using distinct rules for axonal bouton and dendritic spines52, 53 (Fig. 6A). We thus used different η values for bouton and spine growth and applied the same silencing protocol. The network grew properly to the equilibrium state with altered rules but failed to recover synapse numbers (Fig. 6B), only asymmetric connectivity was observed in the input and output synapses of the deprived subpopulation (Fig. 6C). However, a closer look at their network structure revealed the critical role of distance-dependent connectivity patterns in the early work. When distance-dependent connectivity was applied, neurons close to the border of the inhibited area would still receive active inputs from the non-inhibited neighbors. Therefore, these neurons were less deprived than neurons close to the center, even though their external inputs were equally removed.

Accordingly, we were inspired to examine the recurrent inputs’ role in connectivity recovery. We applied the silencing protocol and delivered external stimulation to the deprived subnetwork. In Fig. 6E (Protocol 1), the network in the left panel was simulated at a ten times faster speed than in the middle panels, such that the recurrent connectivity at the timing of stimulation was much lower in the left than in the middle panel. As stated, network dynamics are determined by the product of external inputs and internal connectivity, such that the triggered activity in the deprived subnetwork remained silent in the left panel but was reactivated in the middle panel, despite the same stimulation strength. The connectivity trace in Fig. 6F also confirmed that the residual recurrent connectivity amplified the external stimulation and initiated synapse regeneration and rewiring. However, if the external stimulation were too high (protocol 2), the over-amplification effects of the recurrent connectivity would push the system from silencing-induced degeneration to over-excitation-induced homeostatic degeneration (Fig. 6E,F, right panels). Our simulation results suggested that using recurrent connectivity, external stimulation could modulate neural activity and shape synapse number and network connectivity.

Synaptic scaling is redundant and complementary to biphasic Gaussian rule in maintaining firing rate homeostasis

Following the idea of activity manipulation, homeostatic synaptic scaling could, in theory, sculpt recur-rent connectivity by continuously adjusting functional synaptic transmission during input deprivation. Particularly, homeostatic strengthening of excitatory synapses could increase neural activity and thus may push the activity-dependent structural plasticity from silencing-induced synapse loss to rewiring, as in the case of external stimulation.

To examine this hypothesis, we implemented a monotonic rule for calcium-based synaptic scaling. We grew the network with the biphasic Gaussian structural rule to the equilibrium state and turned on the calcium-based synaptic scaling at the same time when the silencing protocol was initiated (Fig. 7A). To specify the contribution of structural plasticity and synaptic scaling, two types of connectivity were analyzed: (1) structural connectivity (Γstruc.), which was calculated by synapse numbers only; and (2) effective connectivity (Γeffec.), which was calculated by multiplying synapse numbers with the weight of individual synapses. In the scenario where synaptic scaling was absent, the synaptic weights were uniform. Thus, the two types of connectivity were identical (Fig. 7C, left panels). When allowing for weak scaling (Fig. 7B,C, ρ = 0.01), the effective connectivity to the deprived neurons (E-S) increased over time, while its impact on firing rate was negligible and therefore failed to restore synapse numbers and network connectivity. By increasing the scaling strength in the right panel (ρ = 0.02), we observed a higher temporal increment of effective connectivity in E-S synapses which managed to reactive the deprived neurons from silent status and therefore re-initiated the synapse regeneration and rewiring.

Homeostatic synaptic scaling sculpted effective connectivity to interfere with struc-tural plasticity.

(A) Protocol of silencing and synaptic scaling enabling. Three different scaling strengths were applied, that ρ = 0 is for w/o scaling, ρ = 0.01 and ρ = 0.02 represent weak and strong scaling, respectively. (B-C) Time courses of network activity and connectivity. Γstruc. is synapse-number-based structural connectivity. Γeffec. denotes the effective connectivity which multiplies synapse numbers and synaptic weights. (Di-Dii) Structural and effective connectivity matrices of the whole network at t2. (E) Time courses of firing activity and average synaptic weights for an example active excitatory neuron (blue) and an example silent excitatory neuron (grey). In the weight plot, synapses from other silent neurons (solid grey line, S) and from other active excitatory neurons (dashed grey line, E) are specified. (E) Raster plots of 100 selected inhibitory neurons (red), excitatory neurons (blue), and silent neurons (grey) from the network at t1, t2, and t3 labeled in panel B.

We further plotted the structural connectivity matrix (Fig. 7Di) and the effective connectivity ma-trix (Fig. 7Dii) at the time point t2. We showed that although the synapse numbers to the deprived subgroup decreased, the remaining synapses which received inputs from the non-deprived neurons gained weight. This observation was confirmed by the temporal evolution of the mean synaptic weights of two example neurons being deprived (gray) or non-deprived (blue) in Fig. 7E. Our simulation results con-firmed that structural plasticity and synaptic scaling react oppositely but complement each other under extreme activity deprivation. With proper scaling strength, synaptic scaling could initiate synapse regen-eration by carving recurrent connectivity to make the maximum use of recurrent inputs from neighbor neurons. On the other hand, the two rules behaved redundantly under stimulation as expected, that homeostatic disconnection occurred less when homeostatic synaptic down-scaling took place at the same time (Supplementary Figure 4).

While proper scaling strength could restore the synapse numbers and neural activity after silencing, we wondered whether the temporal network dynamics were rescued after synapse regeneration. We sampled three distinct time points (before silencing, during silencing, and after regeneration) indicated in Fig. 7B and plotted the raster plots of inhibitory neurons (red, I), non-deprived excitatory neurons (blue, E), and deprived subpopulation (grey, S). As shown in Fig. 7F, the neurons’ spiking activity was asynchronous and irregular before deprivation (t1), and the deprived neurons became silent as expected during silencing (t2). Intriguingly, synaptic scaling rescued the average firing rate of the deprived neurons, but their activity pattern became highly synchronized at t3. The highly synchronized activity observed may be attributed to the potentiation of the residual active synapses, which dominated the inputs afterward. Our results implied that although the interaction between structural plasticity and synaptic scaling triggered natural synapse rewiring after silencing or denervation, non-homogeneous connectivity among the deprived neurons may introduce unwanted dynamics, which could be of interest in the context of post-traumatic, i.e., injury-triggered-epilepsy.

Hybrid combinations of synaptic scaling and structural plasticity resolve in-consistent results of spine density

Simulations of point neuron networks allowed us to treat individual synapses homogeneously regardless of the neural morphology. However, in a neuron with highly compartmentalized dendritic segments, we may encounter different local combinations of synaptic scaling strengths and structural plasticity time scales in different neural types, which we hypothesize contribute to the inconsistent results of observed spine density in experiments (Fig. 8A).

Systematic study of the interaction between synaptic scaling and structural plas-ticity in response to activity silencing.

(A) We hypothesize that different combinations of synaptic scaling (SS) and homeostatic structural plasticity (HSP) may apply to different neuron types or dendritic segments within the same neuron, or the same type of neuron under different experimental conditions, such that the empirically observed structural plasticity was highly heterogeneous. (Sample neurons were reconstructed based on CA1 pyramidal neuron and dentate gyrus granule cells previously recorded in our lab.) (B) Simulation protocol. We systematically changed the values for the growth rate of the HSP rule (ν) and the scaling strength of the SS rule. (C-D) Example traces of neural activity and struc-tural plasticity of the deprived subpopulation (S neurons) under different parameter combinations. (E-F) Connectivity discrepancies and firing rate discrepancies when different growth rates and scaling strengths were combined. All the discrepancies were calculated by estimating the area between the actual time course and the equilibrium connection probability (10%) or the target rate (ɛ) from the time of silencing until the end of the simulation.

To address this issue, we combined various structural plasticity growth rates (ν) with various synaptic scaling strengths (ρ) in different simulations (Fig. 8B). Example traces in Fig. 8C and D display different time courses of neural activity and structural connectivity of the deprived subpopulation. It varies from condition to condition. To obtain an overview of the impact of these parameter combinations on network rewiring, we first summated the discrepancy between the actual structural connectivity among the subnetwork and its equilibrium state value (10%) over all time points upon silencing until the end of the simulation in Fig. 8E. Cold colors indicate failure in connectivity restoration, and warm colors suggest success in rewiring or over-reconnection. Slower growth rates result in slower spine loss upon silencing and are equivalent to more residual synapses. Clearly, we saw that for slow growth rates, even weak synaptic scaling was sufficient to rescue the synapse loss; for fast growth rates, the small number of residual synapses would require much stronger synaptic scaling to achieve a similar level of synapse restoration.

Similarly, we summated the discrepancy between the actual neural activity and the target rate (7.9 Hz) for the deprived population over all time points since silencing to overview the recovery of neural activity in Fig. 8F. We saw that the combinations that failed to restore network connectivity in Fig. 7E presented negative discrepancies from the target rate (dark blue in Fig. 7F), suggesting a failure in reactivating the silent neurons. However, it is worth noticing that the combinations that succeeded in synapse regeneration in Fig. 7E (warm colors) did not necessarily achieve smooth firing rate restoration. Instead, only those parameters that achieved mild synapse regeneration (light pinkish colors panel E) presented smooth firing rate restoration. Fast growth rates of structural plasticity in combination with strong synaptic scaling strength may lead to very high neural activity (bright colors) which resembles epileptiform activity or result in a drop in the transient firing rate (dark blue colors). Our systematic study confirmed that calcium-based synaptic scaling and homeostatic structural plasticity interact with each other at various time scales and strengths, which may explain the inconsistent experimental results reported.37 Our results also suggested that fast-time-scaled homeostatic structural plasticity and strong synaptic scaling would harm rather than facilitate firing rate homeostasis.

Discussion

The mammalian brain is a complex system consisting of billions of neurons and non-neuron cells packed into layers, regions, and circuits.54 This system is highly robust such that ample external information does not cause catastrophic forgetting,55 nor will input deprivation destroy the network dynamics.7–9 The most extensively observed synaptic-weight-based plasticity in experiments, Hebbian plasticity and homeostatic synaptic scaling, are regarded as major components in balancing flexibility and stability.56, 57 Spine-number-based structural plasticity, however, remains less understood due to divergent experimental observations.58 Nevertheless, structural plasticity, which allows for synapse formation and rewiring, has been proven to greatly increase the memory capacity of the network36 and maintain its criticality,59, 60 suggesting its discernible role in network function and potential interaction with functional plasticity. This manuscript elaborated on the synapse-number-based homeostatic structural plasticity rule and its interaction with the synaptic-weight-based homeostatic synaptic scaling rule. Our time-lapse-imaging experiments revealed a non-linear dependency between spine density and neural activity, which informed a biphasic structural plasticity rule for the systematic study. Computer simulations showed that this biphasic rule and homeostatic synaptic scaling are redundant and interchangeable upon stimulation. However, they behave oppositely under silencing: neurons would reduce spine numbers while potentiating the remaining input synapses. Synaptic scaling may rescue silencing-induced spine loss by restoring neural activity. Our study reconciled varied experimental results and suggested that the redundancy and heterogeneity between homeostatic structural plasticity and synaptic scaling achieve robustness of the complex network of neurons—firing rate homeostasis.

Our work generally extended the scope of structural plasticity by exploring its functional interaction with synaptic scaling. Structural plasticity is less extensively modeled than functional plasticity, due to its high-dimensional morphological data. Many morphological changes may occur in neurons, ranging from dendritic tree branching pattern, axon length, and bouton size, to spine density and head volume, in response to environment and experience, disease and injury, or activity perturbation. These changes would influence network connectivity. Therefore, structural plasticity was frequently proposed to account for functional cortical rearrangement in humans after injury, amputation, or disease-related cognition decay, but only a few dimensions of neural morphology were modeled mathematically. Except for a few models about morphological development,61–63 most structural plasticity models focus on synapse formation and rewiring, which was regarded as the key difference from functional plasticity models. However, early models inherited the synapse-specific view of Hebbian plasticity and modeled the growth and removal of individual synapses. In these models, less-frequently-used or small-weighted synapses were pruned from a highly-connected network, or from a network with regular random synapse growth.64–68 They are variants of the correlation-based rule and also share Hebbian plasticity’s properties and limitations. Unlike synapse-specific correlation-based rules, Butz and van Ooyen proposed a cell-autonomous homeostatic structural plasticity rule that follows a setpoint firing rate or calcium concentration to create or delete spines and boutons randomly.52, 69 This spine-number-based rule is a structural analog to synaptic scaling in maintaining firing rate homeostasis. Apart from that, it preserves the correlation-based associative properties11, 41, 70 and allows for self-organized criticality in developing neural networks,59 which is beyond synaptic scaling and suggests an alternative solution to integrate Hebbian and homeostatic plasticity. Our work used the same model, but added biological features inspired by experiments and investigated its function in more biological scenarios where other plasticity rules also apply.

Particularly, we suggested a biphasic activity dependency for homeostatic structural plasticity with experi-ments. As reported in a recent review,37 spine density alterations observed in experiments seem not as homeostatic as synaptic scaling, especially under prolonged input deprivation. Heterogeneous experimental paradigms may cause inconsistency. Some used lesion methods, for instance, which could increase the secretion of neuroin-flammatory cytokines besides input deprivation.46, 47, 71, 72 Neuroinflammatory cytokines, such as TNF-α, have been shown to influence plasticity induction.73–75 Non-traumatic deprivation methods, such as dark rearing76 or whisker trimming,77 avoided the confounding effects of injury-related cytokine release. But the resulting input deprivation was non-homogeneous on the whole dendritic tree, as apical and basal dendrites usually receive dis-tinct inputs. The same principle may also apply to pharmacological treatments targeting pre-synaptic parts78 or glutamate uncaging in the neighborhood of a few spines.79 However, in studies where drugs homogeneously treated postsynaptic neurons, changes in spine density remain inconsistent.80–85 A closer look at their experi-mental systems suggested a potential role of deprivation strength. Using non-competitive or competitive drugs (with different concentrations), conducting in vivo or in vitro, and embedding within intact circuits or not all lead to the actual magnitudes of inhibition on the target neuron varying from experiment to experiment. Our recent work has demonstrated that depriving synaptic inputs to CA1 pyramidal neurons by lesioning entorhinal cortex or CA3 Schaffer collateral triggered distinct pathway-dependent cellular, structural, and molecular alterations, where the differentiated dominance of the two pathways may play a role among other factors.86 Given that, we hypothesized a biphasic activity-dependent rule to capture the rich scenarios of spine density alterations, that partial activity suppression shall promote spine growth while complete inhibition induces spine loss. Indeed, our time-lapse imaging results confirmed that partial inhibition (200 nM) increased spine density while complete inhibition (50 µM) reduced spine density, and informed our selection of a biphasic rule. Nevertheless, our results also suggested complex behaviors of spine size upon inhibition, which may influence our interpretation of spine density, and we will discuss this later.

In the shape of a Gaussian growth curve, the biphasic structural plasticity model utterly reconciled the rich scenarios of activity-dependent spine density alterations observed in experiments. As indicated by the growth rule, the unstable setpoint (η > 0) leads to varied results in synapse numbers (equivalent to spine numbers) after activity perturbation: increase, no change, or reduction. The silencing-induced synapse loss is especially beyond the reach of any linear (monophasic) rules. As a derived property, this rule enables phase transition by changing neural activity via manipulating external inputs or intrinsic connectivity. By applying external stimulation at proper strengths and timings, for instance, it is possible to reverse the silencing-induced synapse loss into homeostatic rewiring, which may provide insights for using non-invasive brain stimulation modalities to prevent stroke-induced neural lesions.87 However, Butz and van Ooyen have suggested that the Gaussian rule could achieve a natural transition from silencing-induced spine loss to rewiring if different rules for axonal boutons and dendritic spines are used accordingly,52, 53, 88 with which our simulation in Brunel network disagreed. Instead, we showed that the “recovery” observed in their study was the combined effect of distance-dependent connectivity and distinct axonal/dendritic rules. The distance-dependent connectivity created an inhibition gradient, that the neurons at the border were less inhibited than those close to the center, and triggered a gradual connectivity recovery with the Gaussian rule. Our systematic study demonstrated the fundamental effects of neural activity and resulting calcium concentration in regulating retracting and rewiring, regardless of whether it is driven by intrinsic connectivity or external inputs.

Homeostatic structural plasticity usually happens in tandem with functional synaptic scaling, such that they may influence each other’s behavior and make it hard to interpret the observed experimental results. In our experiments, we observed increased spine size and density after three days of 200 nM NBQX treatment. Spine size enlargement may raise the chance of small spines being detected as a spine. Therefore, the spine density increase observed in imaging experiments is a compound effect of spinogenesis, spine stability, and small spines’ potentiation. The same confounding effects happen to interpret synaptic scaling results. Turrigiano et al.22 observed changes in the amplitude but not the frequency of mEPSC after chronic inhibition and excitation, leading to the conclusion of weight scaling, while subsequent experiments reported either frequency and/or amplitude changes. The conversion of silent synapses into active ones89 may increase the frequency but reduce the average amplitude of mEPSCs or sEPSCs due to their small weights. Such changes, by definition, are synapse-number-based structural plasticity, but may not be reflected by imaging techniques, depending on the laser intensity or exposure time. Furthermore, we observed divergent size changes of small and large spines upon complete inhibition. Assuming that shrunk spines may disappear and enlarged big spines will contribute to increased sEPSC amplitudes, we would again obtain different and even opposite results from imaging experiments. Therefore, spine numbers and synaptic weights might be the two sides of the coin named “effective transmission”, and judging it by either side will hide the whole picture of how structural and functional plasticity interfere. Our systematic study tackled this question and demonstrated that the interaction between biphasic structural plasticity and monotonic synaptic scaling achieves an economical and robust control of firing rate homeostasis.

Synaptic scaling originally describes the compensatory changes of synaptic weights in response to activity perturbation, and several variants of mathematical rules were extrapolated and implemented. These models used target firing rate,90–93 normalized synaptic weights,94 or regulated presynapses95, 96 to realize synaptic scaling, except one suggested an integral feedback control algorithm,97 where the authors used a variable with an expo-nential kernel to represent the internal status of neural spiking activity through time. This variable regulates synaptic weights accordingly to reach a target value. However, by the time they came up with the model, it was yet unknown that calcium concentration is the variable that tracks the internal trace of spiking activity.98 Taking advantage of this model, we implemented the calcium-based synaptic scaling rule and studied its interaction with the calcium-based structural plasticity model. Our results first suggested their redundancy. Assuming both rules are monophasic activity-dependent, they are fully redundant, and the one with a faster time scale would dominate. In other words, if synaptic scaling acts faster than structural plasticity, we shall easily observe mEPSC amplitude (or frequency) changes rather than altered spine density. Structural plasticity will join the game only if the counteracting effects of synaptic scaling are insufficient. However, supposing structural plasticity follows a biphasic rule while synaptic scaling follows a monotonic rule, they are only redundant upon stimulation and weak deprivation but bifurcate upon strong inactivity, as confirmed by our simulation. As a result, we may observe neurons with reduced spine density and enlarged spine head sizes in our experiments and others.82, 83 From an economical point of view, spine growth and synapse formation are constrained by the availability of proteins; the transcription and expression of proteins require the energy currency ATPs. Therefore, we postulate that it is more effective for neurons to potentiate the existing synapses than searching for new inputs with new spines. Our sys-tematic study revealed an interdependency between synaptic scaling and structural plasticity, which may explain divergent results obtained in different brain regions, neurons, dendritic segments, and experimental settings.

In summary, how homeostatic plasticity sculpts network connectivity, including synaptic weights and synapse numbers, resembles an integral feedback controller (Supplementary Figure 5), which is ubiquitous in engineering to achieve robust performance regardless of external perturbation. In our setting, both structural plasticity and synaptic scaling rules use calcium concentration as the control signal to drive connectivity kernel modification via negative feedback. Calcium concentration integrates neural firing rates over time and returns a filtered signal responding to its activity. Using this filtered signal rather than the raw firing rate to control the connectivity modification could address the cumulative temporal effect of error in estimating the firing rate. However, the detailed mechanisms of the two rules are slightly different; the former is closer to a Proportional, Integral, Differential (PID) controller, while the latter is a PI controller. Except for the shared integral (I) component, the proportional (P) component for structural plasticity is the growth rate of the growth rule,40 as it provides a direct modification on the number of synaptic elements created or deleted per time unit based on a magnitude relative to the difference between the expected and real values of calcium concentration. For synaptic scaling, the scaling factor (β) regulates the proportional effects. Finally, the unique derivative (D) component for structural plasticity is given by the steepness of the growth curve, which decides how fast or slow the changes should reach the neuron’s desired activity. The derivative component is critical to ensure a smooth and robust response to stimuli by avoiding overshooting and abrupt oscillations that are not biologically realistic, which is especially relevant in the Gaussian case, as the distance to the target firing rate has different accelerations depending on the position on the firing rate (or calcium concentration) axis. Therefore, although the exact growth rule shape remains unknown for homeostatic structural plasticity, we propose that a biphasic curve with changing slope provides a redundant and heterogeneous structural backup for synaptic scaling to maintain robust firing rate homeostasis. This view also prepared us to anticipate their interaction with other plasticity rules where sub-or supra-threshold calcium dynamics are involved.99, 100

Materials and Methods

Ethics statement

We used mouse pups postnatal at 3 to 5 days (P3-P5) from C57BL/6J (wild-type) and Thy1-eGFP mouse lines to prepare entorhinal-hippocampal tissue cultures in the current study. All animals were kept under a 12 h − 12 h light-dark cycle with food and water provided ad-libitum. One male and one or two female(s) were kept within the same cage for breeding. All animal experiments were approved by the appropriate animal welfare committee and the animal welfare officer of Albert-Ludwigs-University Freiburg, Faculty of Medicine under X-21/01B and X-18/02C. All effort was made to reduce the pain or distress of animals.

Preparation of tissue cultures

Entorhinal-hippocampal tissue cultures were prepared as published before.101, 102 All tissue cultures were cul-tivated for at least 18 days inside the incubator with a humidified atmosphere (5% CO2 at 35 °C to reach an equilibrium status. The incubation medium consists of 50% (v/v) 1× minimum essential media (#21575 − 022, Thermo Fisher, USA), 25% (v/v) 1× basal medium eagle (#41010 − 026, Thermo Fisher, USA), 25% (v/v) heat-inactivated normal horse serum, 25 mM 1 M HEPES buffer solution (#15630 − 056, Gibco), 0.15% (w/v) sodium bicarbonate (#25080 − 060, Gibco), 0.65% (w/v) glucose (#RNBK3082, Sigma), 0.1 mg/ml streptomycin, 100 U/ml penicillin, and 2 mM glutamax (#35050 − 061, Gibco). The incubation medium was renewed 3 times per week. If not stated otherwise, the incubation medium applied to the tissue cultures was always pre-warmed to 35 °C and adjusted around pH = 7.38.

Experimental design

The main objective of the experiments in the current study was to probe whether there is a linear or non-linear dose-dependent regulation of dendritic spine density such that we can implement a spine-number-based structural plasticity rule close to biological reality for further systematic studies. Competitive AMPA receptor antagonist NBQX (2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline) at different concentrations was bath applied to wild-type cultures while recording synaptic transmission at CA1 pyramidal neurons. Two magnitudes of inhibition (“partial” and “complete”) were thus determined by the reduction of amplitudes and frequencies of spontaneous excitatory postsynaptic currents (sEPSCs). Then the same concentration ladders were used to treat Thy1-eGFP cultures for three days, where we tracked individual dendritic segments of CA1 pyramidal neurons before and after the three-day treatment to investigate whether two different magnitudes of AMPA-receptor inhibition resulted in divergent alterations of spine densities and spine sizes.

Whole-cell patch-clamp recordings

To probe the effects of NBQX administration at different concentrations on synaptic transmission, whole-cell patch-clamp recordings were conducted in CA1 pyramidal neurons. Recordings were performed at 35◦C. The bath solution contained (in mM) 126 NaCl, 2.5 KCl, 26 NaHCO3, 1.25 NaH2PO4, 2 CaCl2, 2 MgCl2, and 10 glucose (aCSF) and was continuously oxygenated with carbogen (5% CO2/95% O2). Glass patch pipettes had a tip resistance of 4 − 6 MΩ, filled with the internal solution which contained (in mM) 126 K-gluconate, 10 HEPES, 4 KCl, 4 ATP-Mg, 0.3 GTP − Na2, 10 PO-Creatine, 0.3% (w/v) biocytin. The internal solution was adjusted to pH = 7.25 with KOH and reached 290 mOsm with sucrose). We patched 6 neurons per culture to record the spontaneous excitatory postsynaptic currents (sEPSCs) of CA1 pyramidal neurons in the voltage-clamp mode at a holding potential of −70 mV. Series resistance was monitored before and after each recording and the neuron data was excluded if the series resistance went above 30 MΩ. Each neuron was recorded for 2 min.

NBQX treatment

NBQX (2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline) is a competitive antagonist of AMPA receptors.103 We chose two concentrations 200 nM and 50 µM and delivered by bath treatment to achieve a partial or complete inhibition of AMPA receptor currents. Wild-type cultures were either recorded in normal ACSF or ACSF that contained two different NBQX concentrations. In time-lapse imaging experiments, Thy1-eGFP cultures were treated with 200 nM and 50 µM by adding NBQX (Cat. No. 1044, Tocris Bioscience, Germany) in the incubation medium for three days. Whenever we changed the new medium, fresh NBQX was administrated accordingly.

Tissue fixation and immunohistochemical staining

Recorded cultures were fixed and stained for post hoc inspection. Cultures were fixed by immersing into 4% (w/v) paraformaldehyde (PFA) in 1× phosphate-buffered saline (PBS, 0.1 M, pH = 7.38) for 1 h and transferred into 1× PBS for storage at 4 °C after being washed in 1× PBS. Before staining, all fixed cultures were again washed three times with 1× PBS (3 × 10 min) to remove residual PFA. We incubated the cultures with Streptavidin 488 (1 : 1000, #S32354, Invitrogen, Thermo Fisher, USA) in 1× PBS with 10% (v/v) in normal goat serum and 0.05% (v/v) Triton X-100 at 4 °C overnight. In the next morning, cultures were rinsed with 1× PBS (3 × 10 min) and incubated with DAPI (1 : 2000) in 1× PBS for 20 min. After another 4 washes with 1× PBS (4 × 10 min), we mounted the cultures on glass slides with DAKO anti-fading mounting medium (#S302380 − 2, Agilent) for confocal microscope imaging.

Time-lapse imaging

To inspect whether neural spine densities and spine sizes were altered by the three-day administration of NBQX, we employed time-lapse imaging to follow the same apical dendritic segments of CA1 pyramidal neurons before and after treatment. Live cell imaging was performed at a Zeiss LSM800 microscope with 10× water-immersion (W N-Achroplan 10×/0.3 M27; 420947-9900-000, Carl Zeiss) and 63× water-immersion objectives (W Plan-Apochromat 63×/1,0 M27; 421480-9900-000, Carl Zeiss). Thy1-eGFP tissue cultures where clear CA1 pyramidal neurons could be identified were used in this experiment. Dendritic segments from the radiatum layer were imaged. We imaged individual dendritic segments prior to the NBQX treatment and again after the three-day treatment. During the imaging session, the membrane insert with 4 cultures was placed into a 35 mm petri dish filled with 5 ml incubation medium on a platform which was constantly maintained at 35 °C. We used the same pre-warmed and pH-adjusted incubation medium for imaging procedures.

Experimental data quantification

Spontaneous excitatory postsynaptic currents (sEPSCs) were analyzed using the automated event detection tool from the pClamp11 software package as previously described.104

z-stacked fluorescent images of Thy1-eGFP cultures were projected to create a 2D representation of individual dendritic segments. ImageJ plugin Spine Density Counter105 were used to count spine numbers and measure segment length, which estimates spine density. For the same dendritic segments imaged at different time points, special attention was paid to ensure the same starting and ending points were used. Post hoc visual inspection was applied to ensure the spine detection results were not strongly biased. Both raw spine density and normalized spine density by baseline were used in the analysis.

The same z-projected fluorescent images were used to track individual spines for spine size analysis. To eliminate the bias from drawing and automatic reconstruction, we drew circles manually around the spine to cut it from the dendrite, the spine size was estimated by measuring the signal intensity with an arbitrary unit of the drawn circle. The drawing and measurements were performed with FIJI ImageJ. Both the raw values and normalized values by baseline spine size were used in the analysis. Statistical methods were specified in the individual results section.

Statistical analysis

Dunn’s multiple comparison test was applied for statistical analysis regarding the sEPSC events among the three groups. For spine density analysis, the Wilcoxon test was applied to compare the values of each segment before and after the three-day treatment. If not otherwise stated, “ns” means no significant, “*” means p < 0.05, “**” means p < 0.01, “***” means p < 0.001. For spine size analysis, we first applied the Wilcoxon test for individual spines. To account for data clustering within each segment, linear mixed model (LMM) with spine_size timing was applied to compare the values before and after treatment with the segment ID as group factor as used in.42 The significance of LLM results was judged by whether the confidence interval (CI) crossed zero.

Neuron model and network model

We used the same spiking neuron model and network architecture as described before in41 and in.11, 42 Current-based leaky integrated-and-fire point neuron was used for both excitatory and inhibitory neurons. We build an inhibition-dominated network with 10 000 excitatory neurons and 2 500 inhibitory neurons.106 To simplify the scenario, we only grow the connections within the excitatory population (E-E) with the activity-based structural plasticity rule (see the Structural plasticity rule section below). Each inhibitory neuron was beforehand hard-wired randomly to receive synapses from 10% of the excitatory and inhibitory population. All details and parameters concerning neural and network models can be found in the Supplementary Materials. We performed the network simulations with NEST simulator 2.20.2 and NEST 3.0107 and MPI-based parallel computation. All the model parameters and protocols can be found in Supplementary Tables 1-7.

Structural plasticity rule

We enabled the growth, retraction, and rewiring of synapses among excitatory neurons with the help of structural plasticity rules. By definition, each excitatory neuron has multiple dendritic spines and axonal boutons, which are called synaptic elements. Synapses were formed by randomly matching free compatible synaptic elements. The growth and retraction of synaptic elements, or in other words, the number of synaptic elements, is governed by a growth rule. Three structural plasticity rules were explored in the current study: (i) linear growth rule; (ii) Gaussian growth rule with a zero setpoint and a non-zero setpoint; (iii) Gaussian growth rule with two non-zero setpoints. All three rules are determined by a function of calcium concentration that reflects neural activity, where C(t) is the time evolution of calcium concentration. Calcium concentration decays with a time constant τCa and increases with calcium influx (βCa) upon the emission of an action potential S(t) of the postsynaptic neuron. This operation is performed internally by the NEST simulator and works as a low-pass filtered signal of the spiking activity of the neuron. The growth of synaptic elements is regulated differently depending on the calcium concentration and the shape of the growth rule.

Linear growth rule

The linear rule was first introduced in40 and systematically studied in an inhibitory-dominant neural network.11, 41, 42, 70

The number of synaptic elements (z(t)) is linearly dependent on the calcium concentration, where ν is the growth rate, and ɛ is the target level of calcium concentration. Since calcium concentration reflects the neural activity loyally, the target level also suggests a setpoint of firing rate in the context of a certain neural network. As discussed before, when neurons fire below their target rate, they grow new synaptic elements and form new synapses. On the other hand, they break existing synapses and retract synaptic elements when they fire above the target rate (setpoint).

Gaussian growth rule

The Gaussian rule has a more complex dependency on the calcium concentration when neural activity is too low, as shown in Equation 3. This rule was suggested and explored initially in.52 where , and . In this rule, η and ɛ are two setpoints: ɛ is the stable setpoint as used in the linear rule. When the neuron fires above ɛ, it retracts synaptic elements as in the linear rule. η is another setpoint introduced specifically for the Gaussian rule, determining the regulation manner when the neuron activity drops below ɛ. When the neuron is firing below ɛ but above η, the number of synaptic elements present will undergo homeostatic outgrowth, but when the neuron is firing below η, neurons will break synapses and retract elements. In the case where η = 0, neurons cease to change synaptic elements when their firing rate drops to zero.

Homeostatic synaptic scaling

In order to achieve homeostatic synaptic scaling, we make use of a new synaptic model in NEST called scal-ing_synapse. In this synapse model, the weight of the synapse is regulated by the difference between a homeostatic setpoint and the calcium trace of the postsynaptic neuron, where ρ is the scaling factor, and ɛ is the same target value as used in the homeostatic structural plasticity rule.

Activity perturbation

To examine different activity-induced scenarios on neural network connectivity, we performed systematic activity manipulation to a subnetwork of excitatory neurons (Nsub = 1 000), by changing its Poissonian input from 0% to 200% fold of the original intensity (FOI). All manipulations were performed at 6 000 s when the network has grown to the equilibrium state respectively with three structural plasticity rules. For the Gaussian rule with two non-zero setpoints, we applied damping current injection to the soma to facilitate its growth within the first 4 000 s and the network was simulated for another 2 000 s without any facilitating current.

Quantifying firing rate, network connectivity, and synapse number

Firing rate

Firing rate was calculated by the average spike count over a recording period for individual neurons. We have long interval (5 s) and short interval (1 s). Short intervals were only used within the short time window after activity perturbation to reveal its transient dynamics; otherwise, long intervals were used.

Network connectivity

Two types of connectivity were used in the present study. We used a N × N connectivity matrix (Aij) to represent the recurrent excitatory connections of our network, where columns and rows correspond to pre-and postsynaptic neurons. For structural connectivity, the entry Aij of the matrix represents the total number of synaptic connections from neuron j to neuron i. For effective connectivity, we integrated the synapse number with individual weights for each pair of neurons by summating the total weights. So the entry of the connectivity matrix is the equivalent number of unit synapses, by dividing the sum with a uniform weight 0.1 mV. To average the mean connectivity of the whole network or a subnetwork at any given time t, corresponding columns and rows of the connectivity matrix were selected and averaged by . Synapse numbers were calculated by the sum of the entry in the structural connectivity matrix.

Synapse number

Input and output synapse numbers, also called indegree and outdegree in another context, were calculated by summating the input and output synapse numbers of individual excitatory neurons based on the structural connectivity matrix.

Quantifying the discrepancies in firing rate and connectivity from the target values

To apply a systematic comparison among different combinations of the synaptic scaling strengths and structural plasticity growth rates, we summated the discrepancies in firing rate and connectivity from the target values for the subpopulation over time as an index of activity and connectivity recovery. All the discrepancies were calculated by estimating the area between the actual time course and the equilibrium connection probability (10%) or the target rate (ɛ) from the time of silencing until the end of the simulation. The method was explained in detail in Supplementary Fig. 6.

Supportive Information

The authors declare that they have no conflict of interest.

Supporting information

eLife Assessment

This manuscript makes a valuable contribution to understanding the entanglement of homeostatic structural plasticity and synaptic scaling, yet only homeostasis after activity deprivation is studied in depth. The experimental and computational methods are solid but overall incomplete as the link between them remains qualitative. The conclusions drawn from the results are rather vague and their generality or relevance for other research fields is not made clear.

Reviewer #1 (Public Review):

This manuscript investigates how homeostatic structural plasticity and synaptic scaling act under different levels of activity suppression and how this influences the network dynamics during growth and temporary or persistent silencing. To this end, the authors first use electrophysiology and chronic imaging to investigate the influence of different levels of AMPA-receptor blockade. A smaller level leads to reduced activity and up-regulation of synapse size and number, whereas a complete block abolished activity and decreases spine numbers. Along this line, the choice to block AMPAR is unconventional and needs to be better justified as both investigated homeostatic mechanisms are known to be AMPAR dependent.

Second, this finding is transferred into a mathematical rewiring rule, where spine number shrinks, grows, and shrinks again with increasing activity. It is shown that this rule, in contrast to other, simpler rules (grow, shrink), can grow healthy networks from scratch only if additional stimulation is provided. Continuing with these stable networks, the activity of a sub-network is increased, decreased, or silenced by modulating an external stimulation to the neurons. Whereas both activity and connectivity return to a stable state for small alteration, complete silencing leads to disconnection of the silenced network parts. Recovery from this can be achieved by restoring stimulation before the connectivity has completely decayed or by adding sufficiently fast synaptic scaling, although both cases can lead to unhealthy activity. A more systematic assessment of this interaction between scaling and homeostatic rewiring revealed a minimal timescale ratio that is needed for recovery. This is an important step towards disentangling the necessity of multiple, seemingly redundant mechanisms. Yet, in the simulations, the role of recurrent connectivity versus external inputs should be investigated in more detail in order to ensure the generality of the finding that a recovery of the activity is impossible for the presented rewiring rule without synaptic scaling.

Overall, the combination of experiments and simulations is a promising approach to investigating network self-organization. The gradual blocking of activity is especially valuable to inform mathematical models and distinguish them from alternatives. Here, the simulation results clearly demonstrate that the experimentally informed rule exhibits qualitatively different dynamics including the need for another homeostatic mechanism. However, a better connection between the simulations and experiment two would be desirable. In particular, it is unclear whether the model would actually reproduce the experiment, to which other experiments the model results relate, and which experimentally testable predictions the model makes.

In summary, this manuscript makes a valuable contribution to discerning the mathematical shape of a homeostatic structural plasticity model and understanding the necessity of synaptic scaling in the same network. Both experimental and computational methods are solid and well-described. Yet, both parts could be linked better in order to obtain conclusions with more impact and generality.

Reviewer #2 (Public Review):

This manuscript by Lu et al addresses the understudied interplay between structural and functional changes underlying homeostatic plasticity. Using hippocampal organotypic slice cultures allowing chronic imaging of dendritic spines, the authors showed that partial or complete inhibition of AMPA-type glutamate receptors differentially affects spine density, respectively leading to an increase or decrease of spines. Based on that dataset, they built a model where activity-dependent synapse formation is regulated by a biphasic rule and tested it in stimulation- or deprivation-induced homeostatic plasticity. The model matches experimental data (from the authors and the literature) quite well, and provides a framework within which functional and structural changes coexist to regulate firing rate homeostasis.

While the correlation between changes in AMPAR numbers and in spine number/size has been well characterized during Hebbian plasticity, the situation is much less clear in homeostatic plasticity due to multiple studies yielding diverging results. This manuscript adds new experimental results to the existing data and presents a valuable effort to generate a model that can explain these divergences in a unifying and satisfactory framework.

The model and its successive implantation steps are well presented along a clear thread. However, it would have benefited from having an actual timeline of structural changes throughout the three days of AMPAR inhibition, especially as their experimental model allows it. This would have provided additional information on spine dynamics (especially transient spines) and on the respective timescale of the structural and functional changes, and thus led to a better-informed model.

Additionally, the model would have been strengthened by an experimental dataset with homeostatic plasticity induced by higher activity (e.g. with bicuculline). To the best of my knowledge, there is currently no data on structural plasticity following scaling down, and it is also known that scaling up and down are mediated by different molecular pathways. The extension of the model from scaling up (in response to silencing) to scaling down (in response to increased activity) offers an interesting perspective but may be a bit of a stretch.

Finally, the authors are very specific in their definition and distinction of structural and functional homeostatic plasticity for their model. Structural plasticity is limited to spine density and functional plasticity to synaptic scaling, which allows the authors to discuss the interplay between very distinct "synapse number-based structural plasticity" and "synaptic weight-based synaptic scaling", and appears to bypass the fact that spine size regulates the space available for AMPARs at the synapse and thus synaptic weight. The authors are of course aware of the importance of changes in spine size, as they present some intriguing data showing that spine size is differentially affected by partial or complete inhibition of AMPARs and include the putative role of spine size changes in the discussion. However, spine size does not seem to be taken into account in their network simulations, which present synaptic scaling and structural plasticity as completely distinct processes. While the model still offers interesting insights into the interaction of these processes, it would have benefited from a less stringent distinction; this choice and the reasons behind it should be made more explicit in the manuscript.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

(1) Gap of knowledge:

From the introduction, I got the impression that the manuscript tries to answer the question of whether homeostatic structural plasticity is functionally redundant to synaptic scaling. However, the importance of this question needs to be worked out better. Also, I think it is hard to tackle this question with the shown experiments as one would have to block all other redundant mechanisms and see whether HSP functionally replaces them.

We appreciate the reviewer’s valuable feedback regarding the relationship between homeostatic structural plasticity (HSP) and synaptic scaling. The main objective of our study is indeed to investigate whether structural plasticity is homeostatically regulated, and if so, whether it acts as a redundant or heterogeneous mechanism in relation to synaptic scaling, which is widely recognized as a primary homeostatic process.

In our revised introduction, we have clarified this central question and its significance. Specifically, we explored why experimentally observed changes in spine density, a measure of structural plasticity, do not exhibit the same homeostatic characteristics as changes in spine head size, which reflects synaptic scaling, particularly under conditions of activity blockade.

We hypothesized two key points:

(1) Structural plasticity may not follow a monotonically activity-dependent rule as strictly as synaptic scaling.

(2) The observed changes in spine density may be influenced by the simultaneous modulation of spine size, suggesting that structural plasticity and synaptic scaling interact within the same biological system.

Both hypotheses were tested through a combination of experimental observations and systematic computer simulations. Our conclusions demonstrate that spine-number-based structural plasticity follows a biphasic activity-dependent rule. While it largely overlaps with synaptic scaling under typical conditions, it exhibits heterogeneity under extreme conditions, such as activity silencing. Furthermore, our simulations revealed that both mechanisms can compete and complement each other within neural networks.

We believe that these results offer a nuanced understanding of the interaction between structural plasticity and synaptic scaling, highlighting their redundancy under most conditions but also their heterogeneity under specific circumstances. Blocking all other redundant mechanisms, as suggested, would provide a more reductionist view, which may not capture the complexity and interplay of these processes in a physiological setting. Our approach reflects this complexity, providing insight into how these mechanisms operate together in a naturalistic context.

We have revised the introduction to better convey these points and emphasize the significance of this question for understanding the dynamics of homeostatic regulation in neural networks.

Similarly, the simulations do not really tackle redundancy as, e.g. network growth cannot be achieved by scaling alone.

We appreciate the reviewer’s comment regarding synaptic scaling's limitations in achieving network growth. We would like to clarify that we did not intend to suggest that structural plasticity and synaptic scaling are fully redundant. In fact, it is well established in the literature that structural plasticity plays a dominant role during development, particularly in network growth, which synaptic scaling alone cannot achieve.

The primary objective of our study was to investigate the interaction between structural plasticity and synaptic scaling under conditions of activity perturbation, rather than during network growth or development. To avoid any confusion regarding developmental processes, we chose to grow the network using only structural plasticity in our simulations. Synaptic scaling was then introduced (or not) during the phase of activity deprivation to specifically examine its role in regulating homeostasis under these conditions.

We have revised the corresponding sections of the manuscript to clarify this distinction, and we have ensured that the simulations reflect our focus on activity perturbation rather than network development. This distinction should help readers avoid conflating developmental processes with the specific goals of our study.

Instead, the section on "Integral feedback mechanisms" (L112-129) contains a much better description of the actual goals of the paper than is given in the introduction. Moreover, this section does not seem to include any new results (at least the Ca-dependent structural plasticity and synaptic scaling rules seem to be very common for me). I, therefore, suggest fusing this paragraph in the introduction to obtain a clearer and better understandable gap of knowledge, which is addressed by the paper.

We agree that the "Integral feedback control" section provides key information relevant to both the Introduction and Methodology. It outlines the theoretical framework and serves as a basis for the experimental design.

To better reflect this, we have revised the Introduction to include the gap in knowledge. However, we opted to retain the section in the Results, slightly modified, to set the context for the first experiment.

Along this line, as it seems a central point of the manuscript to distinguish the controller dependencies on Calcium, the different dependencies (working models) should be described in more detail. Also, the description of the inconsistencies of the previous results on HSP can be moved from the discussion (l419-l441) to the introduction.

We have revised the manuscript to place less emphasis on the controller models while retaining the core principles of control theory. The description of the HSP model has been moved to the Introduction, as suggested, while the detailed history remains in the Discussion to maintain the manuscript's consistency.

Systematic text revision: Regarding comment (1), we thank the reviewer for suggesting the text reorganization. We have adjusted several parts in the introduction, M&M section, and results section to increase clarity.

(2) Pharmacological Choice:

It should be discussed why NBQX is used to induce the homeostatic effect instead of TTX. As there are studies showing that it might block homeostatic rewiring (doi.org/10.1073/pnas.0501881102) as well as synaptic scaling (10.1523/JNEUROSCI.3753-08.2009), it seems unclear whether the observed effects are actually corresponding to those in other publications.

The rationale for using NBQX in our experiments, rather than TTX, is detailed in the public response. We selected NBQX based on specific experimental motivations relevant to our study’s objectives, while acknowledging the potential differences in effects compared to other studies.

Local text revision: We added one paragraph in the discussion section to explain the idea better.

(3) Model-Experiment Connection:

The paper combines simulations with experimental work, which is very good. However, in my opinion, the only connection between the two parts is that the experiments suggest a non-monotonic dependency between firing rate and synapse density (i.e. the biphasic dependency). The rest of the experimental results seem to be neglected in the modeling part. It is not even shown that the model reproduces the experiments. Instead, the model is tested in different situations and paradigms (blocking AMPARs in the whole culture vs network growth or silencing a sub-population). I think it would make the paper stronger and more consequential when a reproduction of the experiment by the model is demonstrated (with analogue analyses).

The experimental results serve three main purposes. First, as the reviewer noted, the spine analysis was conducted to inform the biphasic rule. Second, spine size analysis was performed to replicate published findings and confirm our modeling results, showing that activity deprivation leads to fewer synapses with larger sizes or higher weights. Third, the correlation analysis of spine density and size across dendritic segments suggested a hybrid combination of two types of plasticity across different neurons.

While we addressed these aspects in the Results and Discussion sections, the collective presentation in Fig. 2 may have caused some confusion. To improve clarity, we have now split the experimental results, presenting them alongside the relevant modeling data in Fig. 2, Fig. 8, and Fig. 9.

Also, there are a few more mismatches between the experiment and the model that you will want to discuss:

• The size-dependent homeostatic effect (l154ff, Fig2F) is not reflected by the used scaling model.

We revised Fig 8 and the corresponding text to explain how the scaling model reflects such an effect.

• The model assumes reduced Ca levels. Yet, the experimental protocol blocks AMPARs, which are to my knowledge not the primary source of Ca influx, but rather the NMDARs.

The model is based on neural activity, with calcium concentration serving as an internal integral signal of the firing rate, allowing for integral control. While calcium plays a critical role in homeostasis, we caution against drawing a strict correspondence between the model's calcium dynamics and the experimental protocol, as calcium can be sourced from multiple pathways in neurons beyond AMPARs, such as NMDARs, voltage gated calcium channels, and intracellular stores. Also, our recent work demonstrated that under baseline conditions, the majority of AMPARs are not Ca2+ permeable, i.e., GluA2-lacking (Kleidonas et al., 2023)

Improving the calcium dynamics, including secondary calcium release and calcium stores, is part of our future plan to refine the HSP model and address experimental findings that are not fully explained by the current model.

• The model further assumes silencing by input removal, whereas the recurrent connections stay intact. Wouldn't this rather correspond to a deafferentation experiment, where connections to another brain area are cut?

Thank you for pointing at this. The modeling section was not intended to directly replicate the tissue culture experiments but rather to provide insights into a broader range of scenarios, including pharmacological treatments, deafferentation, lesions, and even monocular deprivation.

Systematic text revision: Regarding comment (3), the goal of our modeling work was more than reproducing. To better serve the purposes of experimental results used in the present study, to inform, confirm, and inspire, we have systematically adjusted the layout of experimental and modeling results to link them better.

(4) Is the recurrent component too weak?

Your results show that HSP does not restore activity after silencing (deafferentation), whereas you discuss that earlier models did achieve this by active neighbors in a spatially organized network. However, the silenced neurons in your simulations also receive inputs through the "recurrent" connections from their neighbors (at least shortly after silencing). Therefore, given the recurrent input is strong enough, they should be able to recover in a similar way as the spatially organized ones. As a consequence, I obtained the impression that, in your model networks, activity is strongly driven by external stimulation and less by recurrent connections. I understand that this is important to achieve silencing through removing the Poisson stimulation. Yet, this fact may be responsible for the failure to restore activity such that presented effects are only applicable for networks that are strongly driven by external inputs, but not for strongly recurrent networks, which would severely limit the generality of the results. As a consequence, the paper would benefit from a systematic analysis of the trade-off between recurrent strength and input strength. Maybe, different constant negative currents could be injected in all neurons, such that HSP creates more recurrent synapses in the network.

We appreciate this insight. However, increasing recurrent input strength is beyond the scope of the current study, as it would fundamentally alter the predefined network dynamics of the Brunel network used. As noted in the manuscript, complete isolation or cell death is not always the outcome after input deprivation, lesion, or stroke, which cannot be fully explained by the Gaussian HSP rule alone. Butz and colleagues offered a solution using growth rules that maximized recurrent input, and we recognize the importance of their work.

That said, we approached the issue from a different angle, emphasizing the role of synaptic scaling in recurrence rather than relying solely on recurrent input strength. In biological networks, external inputs may vary, recurrency can be weak or strong, and synaptic scaling can dominate. Our model offers a complementary hypothesis, suggesting that these factors, in combination, contribute to the diverse and sometimes contradictory results found in the literature, rather than posing a strict constraint on network topology.

Local text revision: We emphasized these points in the Discussion section again.

(5) Missing conclusions / experimental predictions

As already described, the modelling work is not reproducing the presented or previous experimental data. Hence, the goal of modelling should be to derive a more general understanding and make experimental predictions. Yet, the conclusions in the discussion stay superficial and vague and there are no specific experimental predictions derived from the model results.

For example, the authors report that the recovery of activity in silenced cultures is observed in a previously spatially structured model but not in theirs -- at least with slow or no scaling. Yet it is left to the reader to think about whether the current model is an improvement to the previous one, how they could be experimentally distinguished, or to which experimental findings they relate or compare, which I would expect at this point. I would advise reworking the discussion and thoroughly working out which new insights the modelling part of the study has generated (not to be confused with the assumptions of the model aka the biphasic plasticity rule) and relating them to experimental pre- and postdiction.

We recognize the reviewer’s concern, which is closely related to comment (4). We have addressed these points by reorganizing the text to better clarify the purpose of our experimental work and its connection to the modeling results.

Specifically, we have reworked the discussion to highlight the new insights gained from the modeling, and how these can inform experimental predictions and interpretations. This includes distinguishing our model from previous ones and providing clearer connections to experimental findings.

Systematic text revision: Most of the comments on combining experiments and modeling results and on developing the story based on our expectations raised here are sincere and may also reflect the expectations and concerns of a broader readership, so we have accordingly adjusted the text in the Results and Discussion sections to make our points clear.

Suggestions for minor changes:

Fig 1I: Please check the graph and make it more self-explaining. For example, mark the "setpoint" activity (in my opinion, both curves should be at baseline there. In that case, however, I do not see the biphasic behavior anymore). Maybe the table and the graph can be aligned along the activity axis? Also: synaptic inhibition should be increased and not decreased, right?

Local text and figure revision: I guess the reviewer meant for Fig. 2I? We have improved the visualization to avoid confusion.

L74-81: I would reverse the order of associative and homeostatic plasticity in this paragraph.

Local text and figure revision: We have fine-tuned the order in the first and second paragraphs to match the readers' expectations.

L74-75: Provide references for such theories.

Local text and figure revision: fixed.

L84-86: Please provide a reference for the claim that negative feedback, redundancy, and heterogeneity contribute to robustness.

Local text and figure revision: fixed.

L 95-97: I think the heterogeneity aspect needs to be worked out a bit better. Do you mean that the described mechanisms contribute to firing rate homeostasis in a different mixture for each neuron (as shown assumed in the last figure)?

Local text and figure revision: The term heterogeneity is used in the manuscript for two major different settings: (1) heterogeneity in terms of control theory and (2) different combinations of HSP and SS rules. We have named the second condition as diversity to avoid confusion.

L 132: The question of linearity has not been posed so far. Also, I think "monotonous" would be a much better term than linear (as a test for linearity would require more than 2 datapoints).

Local text and figure revision: We agreed linear is not a good term. We replaced it with ‘monotonic’ throughout the manuscript.

Fig2 Bii: The data for 50um is clearly not Gaussian.

We did not imply that the 50 µM condition is Gaussian. Instead, we noted that the non-linearity observed in both the 200 nM and 50 µM data suggests a non-monotonic growth rule rather than a linear one. We applied the Gaussian rule because it has been extensively studied in previous simulations, allowing us to benchmark our findings against those results.

Fig2 D, E inset: The point at time 0 does not convey any information and could be left out.

The time zero data is included to demonstrate that the three groups have a similar baseline, ensuring that any observed differences are due to the treatment and not pre-existing biases in the grouping.

L 178: As the Gaussian rule drops below zero above the upper set-point again, it is rather tri-phasic than bi-phasic.

We intended to convey that inhibition results in either spine growth or deletion, reflecting a bi-phasic response rather than a true tri-phasic one.

Fig 6A: You may want to mark the eta variables in the curves.

Local text and figure revision: fixed.

Fig 6E: The curve of the S population extending to the next panel looks a bit messy.

We retained the curve extension to visually convey the impression of excessive network activity.

L272: It needs to be better described/motivated how protocol 1 and 2 are supposed to study the role of recurrent connection as well as what kind of biological situation this may be.

Local text and figure revision: The corresponding text has been adjusted to avoid confusion.

L 272: It is not clear how faster simulation leads to less recurrent connectivity, when the stimulation protocol and the rates stay the same and the algorithm compensates for the timestep properly. Maybe you rather want to say that you silence 10x longer and stimulate 10x longer?

Local text revision: The corresponding text has been adjusted to avoid confusion.

L. 302: "reactivate"?

Local text revision: fixed.

L 322f: I would suggest showing the connectivity matrix for a time-point with restored activity as well.

Local text and figure revision: fixed.

Fig 8A: The use of the morphological reconstructions is a bit misleading as the model uses point neuron.

Local text revision: Now after reorganization, it is in Fig.9. We kept the reconstruction figure for motivational purposes, suggesting how to understand the meaning of the combinations in more biologically realistic scenarios. The corresponding text has been adjusted to avoid confusion.

Fig 8E-F: the y axis should be in the same orientation as in panel D.

Local text and figure revision: Good idea and fixed in the new Fig. 9.

Fig. 8F: The results here look a little bit random. Maybe more runs with the same parameters would smooth out the contours or reveal a phase transition.

Local text and figure revision: Thank you for the suggestion. We conducted an additional ten random trials to average the traces and heatmaps, improving the clarity of the results now presented in Fig. 9.

L411: Note that there are earlier HSP models by Damasch and van Ooyen & van Pelt, that might be worth discussing here.

Local text revision: fixed.

L416 "beyond synaptic scaling" reference needed.

Local text revision: fixed.

L419: The biphasic rule was suggested by Butz already.

Local text revision: We adjusted the text to emphasize our contribution in suggesting/confirming the biphasic rule based on direct experimental observations.

L 419-44: Most of this is actually state-of-the art and may be better placed in the introduction to justify the use of NBQX as a competititve blocker.

Local text revision: We adjusted the text in the introduction and Discussion sections to cover the raised points.

L487: In my opinion, although scaling adapts the weights quickly, the information about deviating firing rate is still stored in the calcium signal such that it will also give rise to structural changes (although they may be small when the rate is low). Thus, I think that fast scaling does not abolish structural changes.

Local text revision: We adjusted the text to account for other factors that could lead to the same or opposite conclusions.

L502f: Sentence unclear. Do you mean Ca is an integrated (low-pass filtered) version of the firing rate?

Yes.

L504: What is the cumulative temporal effect of error in estimating firing rates?

We were referring to the potential instability in numeric simulations if the firing rate is not tracked by an integral signal (calcium concentration) but is instead estimated through average spike counts over time. In our model, calcium serves as a proxy for the firing rate to guide homeostatic structural plasticity. The intake and decay constants are set to minimize the accumulation of errors over time, making long-term error accumulation unlikely. In any case, this is not intended to be a precise measure of the firing rate but rather a smooth guide for homeostatic control.

Local text revision: We rewrote the section so as not to cause extra concerns.

L505: Which two rules are meant here? Ca- and firing rate based or HSP and scaling?

Local text revision: The two rules are the HSP rule and the HSS rule. We have adjusted the text to improve clarity.

L505ff: I did not really understand the control theoretic view here and Supp Fig 5 is not self-explaining enough to help. In my view, scaling is a proportional controller for the calcium level (the setpoint is defined for calcium and not firing rate). Also, all of the HSP rules do neither contain an integral nor a differential of the error and are thus nonlinear but proportional controllers in first approximation. If this part is supposed to stay in the manuscript, the supporting information should contain a more detailed mathematical explanation. Relevant previous work on homeostatic control by synaptic scaling and homeostatic rewiring, e.g. doi: 10.23919/ECC54610.2021.9655157 should be discussed

Local text revision: We have updated the last paragraph to increase clarity. The HSP and HSS rules are proportional and integral for neural activity, as neural firing rate homeostasis is the meaningful goal. However, it is also correct that the integral component is gone if we view calcium concentration as the goal or setpoint. This paper is discussed and cited in a paragraph above this one.

Reviewer #2 (Recommendations For The Authors):

I have some additional suggestions and questions for the authors, which I am presenting following the order of the figures.

Fig 1A: I'm a little bit puzzled by the timescales between Hebbian and homeostatic plasticity; a wealth of data suggests that Hebbian plasticity acts on a faster timescale than homeostatic plasticity, while Aii-Aiii implies the opposite. In lesion-induced degeneration, for instance, which is mentioned later by the authors, spine loss has been suggested to be Hebbian (LTD) while the subsequent recovery is homeostatic. Additionally, it will not be clear to the reader if the same stimulus could induce Hebbian and homeostatic plasticity, or why; the rest of the manuscript seems to imply that any stimulus could and would trigger homeostatic plasticity, which is not the case. Finally, there should be a mention somewhere that Hebbian structural plasticity also exists.

Local text and figure revision: We thank the reviewer for pointing out the time scale issue, which was not explicitly considered here and is now updated.

Fig. 2Bii: There is no significant difference at 200nm NBQX for sEPSC amplitude, contrary to what is stated in the text (line 136). Which one is it?

Local text revision: We thank the reviewer for pointing out the mistake. We have inspected the original statistical file and corrected the text.

Fig. 2F: The description of Fig. 2F in the text confused me for the longest time. I am still unsure why 200nm NBQX is described as leading to a general size increase when it follows the control line so closely, crosses 0 at the same point, and is even below the control line for the largest spine sizes. Similarly, 50um NBQX neatly overlaps with the control condition except for the smallest and largest spines, so the "shrinkage of middle-sized spines" doesn't seem different from the control condition. I also couldn't find any data supporting the statement that 50um NBQX increased only the size of "a small subset of large spines". Maybe the authors could clarify this section? I would also suggest adding statistics between the treatments at each spine size bin to support the claims, as they are central to the rest of the paper.

Importantly, there is no description of the normalization nor the quantification of the difference between days in the methods; I am assuming post-pre for the difference and (post-pre)/pre for the normalization, but this should be much more detailed in the methodology. I was happy to see the baseline raw spine sizes in Supplementary Fig. 1, and would also suggest adding the raw spine sizes after treatment for comparison.

Local text and figure revision: We have adjusted the text and figure to improve clarity.

Fig. 2G/S2A: a scale for the label sizes would be helpful. I would also like to have the same correlation for 50um NBQX treatment and the control condition (at least in the supplementary figures).

Local text and figure revision: We have adjusted the text and figure to improve clarity.

Fig. 2I: I might be missing something, but why is the activity line flat when there are changes in spine density and size?

Local text and figure revision: We have adjusted the text and figure to improve clarity.

Fig. 3C-D: they are referenced in the text as Fig. 1C-D (lines 188-194).

Local text revision: fixed.

Fig. 5: it is interesting that the biphasic model captures both spine loss and recovery, fitting well with lesion-induced degeneration and recovery. Does this mean that the model captures other types of plasticity, or does it suggest to the authors that both steps are homeostatic?

Indeed, the biphasic HSP rule captures two types of activity dependence. The pioneering work by Gallinaro and Rotter (2018) also demonstrated that the HSP rule, even in its monotonic/linear form, exhibits associative properties, which are typically associated with Hebbian plasticity.

Fig. 6A: This figure requires a more detailed legend - what are the various insets? Does the top right graph only have one curve because they are overlapping and the growth rules are the same for axons and dendrites?

Local text revision: fixed.

Fig. 6E: There is usually an overshoot when a stimulus is removed, in this case at the end of the silencing period (as shown in Fig. 1Aiii). Is there a reason why this is not recapitulated here? It shouldn't be as extreme as in the right panel so there should be no degeneration.

We agree that removing the stimulus would typically trigger an opposite homeostatic process. However, in this protocol, we aimed to emphasize the role of recurrency by presenting extreme cases to illustrate potential scenarios for the readers.

Local text revision: We revised this paragraph to walk the readers through the rationale better.

Fig. 6: the authors mention distance-dependent connectivity (line 268), but I couldn't find any data related to that statement. I was particularly curious about that aspect, so I would like to know what this statement is based on, especially as they touch again on the role of morphology in Fig. 8, and distance-dependent connectivity is more prominent in the discussion. On a similar note, would the authors have data from other layers of CA1 that would show similar or other rules? Please note that I am not asking to include these data in the present paper - I am just curious if these data exist (or if the experiments are considered).

Such an extensive dataset is included and thoroughly investigated in another study that has just been published in Lenz et al., 2023. We updated the reference in the revised text.

Fig. 7E top: the scalebar is missing.

Local text revision: fixed.

Fig. 8A: do the colors have meaning? If yes, please state them. Also indicate that the left two neurons are pyramidal cells from CA1 and the right neurons are granule cells from the dentate gyrus.

Local text revision: fixed.

Line 302: "reactive" should be "reactivate".

Local text revision: fixed.

Version 2

Abstract

Critical network states and neural plasticity are essential for flexible behavior in dynamic envi-ronments, allowing for efficient information processing and experience-dependent learning. Synaptic-weight-based Hebbian plasticity and homeostatic synaptic scaling were considered the key mechanisms in enabling memory while stabilizing network dynamics. However, the role of structural plasticity as a homeostatic mechanism is less consistently reported, especially under activity inhibition, leading to an incomplete understanding of its functional impact. In this study, we combined live-cell microscopy of eGPF-labeled neurons in organotypic entorhinal-hippocampal tissue cultures with computational modeling to investigate the response of spine-number-based structural plasticity to activity perturba-tions and its interaction with homeostatic synaptic scaling. Tracking individual dendritic segments, we demonstrated that inhibiting excitatory neurotransmission does not monotonically regulate den-dritic spine density. Specifically, inhibition of AMPA receptors with a low concentration of 2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline (NBQX, 200 nM) significantly increased spine density, while complete AMPA receptors blockade with 50 µM NBQX reduced spine density. Motivated by these findings, we developed network simulations incorporating a bi-phasic structural plasticity rule governing activity-dependent synapse formation. We showed that this biphasic rule maintained neu-ral activity homeostasis under stimulation and permitted either synapse formation or synapse loss, depending on the degree of activity deprivation. Homeostatic synaptic scaling affected the recurrent connectivity, modulated the network activity, and influenced the outcome of structural plasticity. Specifically, it reduced stimulation-triggered synapse loss by downscaling synaptic weights and res-cued silencing-induced synapse loss by upscaling recurrent inputs, thus reactivating silent neurons. Our interaction between these mechanisms offers an explanation for divergent findings in the existing literature. In summary, calcium-based synaptic scaling and homeostatic structural plasticity rules compete and compensate for one another, ensuring efficient and robust control of firing rate home-ostasis.

Introduction

To survive in dynamic environments, animals must quickly respond to familiar cues, such as those signal-ing food or predators, while remaining alert to novel stimuli. The former reflects experience-dependent learning, where subtle memory cues can trigger strong responses in the corresponding pathway during re-call.1–4 The latter requires brain networks to maintain activity at a critical state, allowing information to be transmitted as action potentials and conveyed as neural avalanches.5–10 These processes represent two seemingly opposing but well-coordinated mechanisms: associative learning and firing rate homeostasis. The idea of firing rate homeostasis has been hovering in theories since the discovery of long-term synaptic potentiation (LTP),.11 LTP is a positive-feedback mechanism that adjusts synaptic strength among ex-citatory neurons (“neurons that fire together, wire together”12), as postulated by Donald Hebb.13 While its associative properties allow synapses to preserve memory traces as increased synaptic weights, it also poses the risks of driving network dynamics toward overexcitation or silence14 (Fig. 1Ai,Aii,Bi). In vivo studies in rodents have demonstrated firing rate homeostasis in the visual cortex and hippocampus, where neural activity is restored at both individual neuron15–17 or population level18, 19 within days of pertur-bation. This ability to restore network dynamics without erasing memories highlights the robustness of the complex system—the human brain network.20–25

Overview of Hebbian and homeostatic plasticity.

(Ai) Neural network activity is driven by external inputs in a static network. (Aii) Hebbian plasticity amplifies network responses to external inputs through a positive feedback mechanism. (Aiii) Homeostatic plasticity restores activity to a set-point via a negative feedback mechanism, essential for maintaining firing rate homeostasis. (Bi) Hebbian functional plasticity is synapse-specific, strengthening recurrent connectivity by potentiating the weights of specific synapses involved in neural activation. In contrast to synapse-specific Hebbian potentiation, synaptic depression can be induced in the same activated synapse (homosynaptic depression) or neighbor-ing synapses (heterosynaptic depression) using specific protocols. (Bii) Homeostatic synaptic scaling is cell-autonomous, involving proportional upscaling or downscaling of all input synaptic weights in response to chronic changes in neural activity. (B) Homeostatic structural plasticity is also cell-autonomous char-acterized by compensatory spine loss during chronic excitation of the postsynaptic neuron. Conversely, chronic inhibition induces divergent, often non-homeostatic, changes in spine density.

Robustness is a ubiquitous concept in many complex systems or engineering systems, meaning that when external perturbations forcefully change its dynamics, the system adapts to return to its origi-nal attractor or moves to a new attractor state that preserves its functions.23, 24 Essential features of robustness include negative feedback control, redundancy, and heterogeneity.20–25 For instance, homeo-static synaptic scaling exemplifies negative feedback strategy26 (Fig. 1Aiii,Bii), as neurons proportionally adjust their synaptic weights in a compensatory manner in response to long-term changes in neural ac-tivity,26 membrane potential,27 or intracellular calcium concentration.28 Redundancy and heterogeneity encompass various mechanisms that achieve the same goal, providing backup in case of failure. Al-though redundancy and heterogeneity have been recognized in biological systems in the form of genetic buffering and convergent molecular circuits,23, 24, 29 their representation in activity-dependent plastic-ity has been less explicitly discussed. Induction of homosynaptic long-term depression (LTD)30 under the Bienenstock–Cooper–Munro (BCM) rule31 or spike-timing-dependent plasticity (STDP) rule,32–35 or induction of heterosynaptic LTD via synaptic tagging and capture36 contribute to stabilizing firing rates. Inhibitory plasticity also helps restore the balance between excitation and inhibition.37–39 These mechanisms, which modulate synaptic transmission locally or globally, exhibit redundancy in function and heterogeneity in the implementation in reference to homeostatic synaptic scaling, serving the same purpose of stabilizing the network dynamics.

However, it remains unclear whether such redundancy extends to the structural substrate of synaptic transmission. Structural plasticity encompasses changes in sizes or numbers of dendritic spines and axonal boutons, synapse numbers, and network connectivity, all of which are critical to functional transmission and memory capacity.40 The homeostatic structural plasticity (HSP) model, which posits that synapse formation and elimination are homeostatically regulated,,41–45 has expanded the potential for modelling network reorganization in response to diverse perturbations.19, 46–49 Nonetheless, experimental results present a more complex narrative (Fig. 1Biii): Studies often find spine loss during synaptic downscaling, while inconsistent and mostly non-homeostatic spine loss is seen in upscaling experiments under chronic activity deprivation.50 This complexity underscores the need for a deeper understanding of activity-dependent structural plasticity, particularly the rules governing spine number changes. Given the range of experimental conditions, including monocular deprivation, lesions, and pharmacological treatments, we hypothesized that: (i) the relationship between neural activity and spine numbers is non-monotonic, and (ii) homeostatic synaptic scaling might confound the interpretation of structural plasticity rules based on the inconsistent experimental results.

In this study, we used two concentrations of the competitive AMPA-receptor antagonists NBQX to gradually block neural activity. Using whole-cell patch-clamp recordings and time-lapse imaging, we found that activity deprivation affects dendritic spine numbers and sizes in a non-monotonic manner. A low concentration of NBQX (200 nM) increased spine numbers, while a higher concentration (50 µM) reduced them. These findings guided us to establish a spiking neural network model with a biphasic, synapse-number-based HSP rule, which successfully captured both homeostatic and non-homeostatic changes in response to varying degrees of activity perturbations. Additionally, we implemented a mono-tonic, synaptic-weight-based homeostatic synaptic scaling (HSS) rule based on the literature.51 Although largely redundant to the biphasic HSP rule, the HSS rule facilitates a shift from non-homeostatic synapse loss to homeostatic synapse regeneration by adjusting recurrent connectivity in silenced networks. Our results highlight the redundancy and complementarity between these two rules, underlining the combined importance of HSP and HSS in enabling robust and efficient adaptation of network activity.25, 52, 53

Results

Informing a non-monotonic structural plasticity rule with gradual activity blockade experiments

The variability in experimental results regarding structural plasticity under chronic activity depriva-tion50 could arise from either non-monotonic, activity-dependent regulation of spine numbers, concurrent changes in synaptic weights and firing rates due to synaptic scaling, or a combination of both. To com-prehensively address this, we established computer simulations that integrate the synapse-number-based structural plasticity rule with the synaptic-weight-based synaptic scaling rule. These activity-dependent models function as integral feedback mechanisms linked by the same intracellular calcium concentration C(t) and setpoint value ɛ (Fig. 2A), which monitors neural activity and influences synaptic weights or synapse numbers. This approach eliminated the need for manual time-scale alignment. Structural plas-ticity models vary, with some suggesting neural activity affects neurite growth and retraction, thereby influencing synapse formation and loss either monotonically19, 46, 47 or non-monotonically.41, 43–45 Given the need for additional experimental data to inform model selection, we conducted experiments to test whether spine density responds monotonically to changes in neural activity.

Non-monotonical modification of spine numbers following three-day synaptic inhibi-tion via NBQX informs a biphasic structural plasticity rule.

(A) The integral feedback control framework for homeostatic synaptic scaling and structural plasticity. Both rules use intracellular calcium concentration ([Ca2+]i, C(t)) to track neural activity (AP, action potential, S(t)). Calcium concentra-tion is updated each time via calcium influx (βCa) upon the emission of a post-synaptic action potential and decays with a time constant τCa. Synaptic scaling follows a weight-dependent rule, multiplicatively adjusting the synaptic weight w(t) with a scaling factor ρ, based on the discrepancy from the setpoint ɛ of calcium concentration. Structural plasticity uses the same guidance signal (i.e., calcium concentration) to regulate the growth and retraction of axonal boutons and dendritic spines, aiming to reach the same setpoint value. Two versions of the structural plasticity rule are shown: one with a linear dependency and the other with a non-linear dependency on intracellular calcium concentration. (B) Example CA1 pyramidal neuron recorded from entorhinal-hippocampal tissue culture to assess the effects of different NBQX concentrations. Scale bar: 500 µm. (Ci-Cii) Group data of spontaneous excitatory postsynaptic currents (sEPSC) in three groups (N = 18 for the control group, N = 18 for the 200 nM NBQX-treated group, N = 12 for the 50 µM NBQX-treated group). (D) Example Thy1-eGPF culture and dendritic segments from the stratum radiatum (rad.) before and after three-day treatment. Scale bars: 200 µm and 5 µm. (E) Spine density at baseline and after the three-day treatment. All values are normalized to the corresponding baseline values. Light-shaded lines show raw data (solid lines represent increased spine density; dashed lines represent decreased density). Dark-shaded lines with error bars represent the means and standard error of the means (s.e.m.s) for each group. (N = 19 for the control group, N = 24 for the 200 nM-treated group, N = 33 for the 50 µM-treated group.)

We used two concentrations of the AMPA receptor antagonist NBQX – 200 nM and 50 µM – and studied CA1 pyramidal neurons in organotypic entorhinal-hippocampal tissue cultures (Fig. 2B). Whole-cell patch-clamp recording of AMPA receptor-mediated spontaneous excitatory postsynaptic currents (sEPSCs) revealed a significant reduction in mean sEPSC amplitude with 200 nM NBQX and a near-to-complete blockade of excitatory synaptic transmission with 50 µM NBQX (Fig. 2Ci-Cii). Both NBQX concentrations were applied to Thy1-eGFP cultures for three days to induce chronic activity inhibition. Individual dendritic segments (side branches of apical dendrites in the stratum radiatum) were imaged before and after treatment using time-lapse microscopy (Fig. 2D). The analysis showed that 200 nM NBQX increased spine density (p = 0.003, Wilcoxon test), while 50 µM NBQX reduced spine density (p = 0.008, Wilcoxon test), compared to their baseline densities (Fig. 2E). No significant changes in spine density were observed in the control cultures (p = 0.06, Wilcoxon test). These results demonstrated a non-monotonic relationship between neural activity and spine numbers under activity deprivation: partial inhibition increased spine numbers, while complete inhibition led to a reduction.

Stabilizing and characterizing a biphasic structural plasticity rule

Since changes in spine numbers or density are functionally linked to synapse formation, loss, and rewiring, we aimed to establish structural plasticity rules in point neurons and study synaptic rewiring in a large, topology-free spiking neuron network, thereby bypassing the complexities of neural morphology (Fig. 3A). For comparison, we first simulated the linear growth rule (Fig. 3Bi). To capture the observed non-monotonic dependency in a simplified manner, we adopted a Gaussian-shaped growth rule for synaptic element numbers with two setpoints. By setting the first setpoint at either η = 0 or η > 0, we generated two variants of the Gaussian rule (Fig. 3Bii-Biii). We then grew three neural networks where the synapses among excitatory neurons were governed by these three rules, respectively (Fig. 3C). As reported in previous studies,19, 45, 46, 48 the linear-rule-guided network developed smoothly to a homogeneous sparsely connected network (10% of connection probability, equivalent to approximately 1 000 excitatory synapses per neuron) with an average firing rate near the target value of ɛ = 7.9 Hz (Fig. 3Di). The Gaussian rule with a zero setpoint (η = 0) also developed the network to an equilibrium state similar to the linear rule (Fig. 3Dii). However, the Gaussian rule with two non-zero setpoints (η = 0.7 and ɛ = 7.9) failed to develop a proper network (Fig. 3Diii).

Growing a neural network with three distinct structural plasticity rules.

(A) A network of point neurons was used to study structural plasticity, with simplified dendritic morphology. Dendritic spines are depicted as pink sticks on the soma, and axonal boutons are represented by empty or solid half-circles. An empty half-circle with a dashed line indicates an axon during retraction. In this model, intracellular calcium concentration is linearly correlated with neural firing rate, so neural activity is used to reflect “firing rate homeostasis” throughout the manuscript, as it has a meaningful physical unit compared to the arbitrary unit (a.u.) for calcium concentration. (Bi-Biii) Three structural plasticity growth rules regulate changes in synaptic element numbers. (Bi) Linear rule with a single setpoint (ɛ = 7.9, orange line). (Bii) Gaussian rule with two setpoints, one at zero (η = 0, grey line; ɛ = 7.9, orange line). (Biii) Gaussian rule with two non-zero setpoints (η = 0.7, grey line; ɛ = 7.9, orange line). Three shades represent 100%, 50%, or 10% of the original growth rate (ν), with positive and negative values indicating the rate of synaptic element growth or loss. (C) Neural network architecture based on the Brunel network model, consisting of 10 000 excitatory (blue, E) and 2 500 inhibitory neurons (red, I) stimulated by external Poissonian inputs (P). All I-I, I-E, and E-I synapses are hard-wired with 10% probability, while E-E synapses are subject to structural plasticity rules. (Di-Diii) Temporal dynamics of neural activity and network connectivity (Γ) during network growth, guided by the three distinct rules. Unless otherwise stated, the curve and shaded areas in activity plots represent the mean and standard deviation of the neural activity for the I and E populations. The network reached an equilibrium state (Γ = 10%) in Di and Dii but not in Diii. The firing rates distribution and network connectivity matrices of the indicated time points (solid triangles) are provided in Supplementary Fig. 3 for Di-Dii, and in Fig. 4 for Diii.

Closer examination of the neural firing rates and network connectivity at the late growth stage of this network (Fig. 3Diii) revealed that half of the excitatory neurons were silent (Fig. 4A), and the network exhibited inhomogeneous connectivity, with nearly half of the neurons isolated (Fig. 4Bi-Bii). Correlating the firing rates with synapse numbers of individual excitatory neurons confirmed that the silent neurons were isolated, while neurons with firing rates close to the target possessed approximately 1 000 synapses, similar to the linear rule (Fig. 4C). These results align with the properties of the Gaussian growth rules: one setpoint is stable (ɛ = 7.9), while the second setpoint can be stable if the synapse numbers remain unchanged after silencing (η = 0) or unstable if disconnection occurs after silencing (η > 0). We refer to the Gaussian rule with two stable setpoints as stable Gaussian rule, while the variant with one unstable setpoint is termed the biphasic Gaussian rule in this manuscript.

Silent neurons remain isolated in the network regulated by the biphasic Gaussian rule.

(A) Histogram of firing rates for excitatory and inhibitory neurons sampled at the time point indicated in Fig. 3Diii. Nearly half of the excitatory neurons were silent. The blue vertical line repre-sents the mean firing rate of non-silent neurons, while the orange dashed line marks the target firing rate (ɛ = 7.9 Hz). (Bi-Bii) Network connectivity matrix and distribution of synapse numbers for individ-ual excitatory neurons. (C) Correlation heatmap showing the relationship between neural activity and synapse number of individual excitatory neurons. Silent neurons did not form synapses, whereas active neurons firing around the target rate formed approximately 1 000 synapses with other active excitatory neurons. (D) Network architecture when a damping facilitating current (Ifacilitating) was injected to en-hance network development. (E) Temporal dynamics of neural activity and network connectivity when a facilitating current was injected. The inset shows the firing rate distributions of both excitatory and inhibitory neurons at a selected time point (solid triangle). The facilitating current started at 750 pA and decayed linearly to zero during a 4 000 s period. (F) Different initial values of facilitating currents resulted in varying network connectivities. An initial value of 750 pA was used throughout the manuscript.

Inspired by previous observation that increased neural excitability facilitates circuit development,54 we enhanced the excitability of excitatory neurons during network growth by applying a damping facilitating current (Ifacilitating) to raise the average membrane potential closer to the threshold potential (Fig. 4D). With this intervention, the neural network governed by the biphasic Gaussian rule successfully grew to the equilibrium state and maintained the dynamics after the facilitating current decayed to zero (Fig. 4E). Given that varying the initial intensity of the facilitating current resulted in different equilibrium connectivity (Fig. 4F), we used the intensity (750 pA) that produced the same network connectivity (10%) and firing dynamics as the linear rule for further exploration.

The biphasic Gaussian rule reconciles homeostatic and non-homeostatic struc-tural changes triggered by activity perturbation

As a proof-of-concept, we systematically modulated the input strength to a subgroup of excitatory neu-rons (10% of the entire excitatory population) after the growth period. Input strength was varied from 0% to 200% fold of the original intensity (FOI, Fig. 5A), mimicking both input deprivation and stim-ulation. Fig. 5Bi-Biii,C illustrates three example protocols demonstrating neural activity and network connectivity under the biphasic Gaussian rule. Intuitively, stimulation increased the activity of the tar-geted subpopulation (S) (left panel in Bii) and triggered a reduction in connectivity both among the stimulated neurons (S-S) and between the stimulated and non-stimulated excitatory neurons (S-E; left panel in Biii). The connectivity matrices in panel C display the final network connectivity. Ultimately, homeostatic disconnection restored the activity levels of the stimulated neurons to the setpoint value (orange line in Bii). In contrast, the outcome of activity deprivation depended on its severity. Both mild deprivation and silencing reduced the neural activity of the affected subpopulation (middle and right panels in Bii). However, only mild deprivation allowed the network to restore activity to the setpoint value via a homeostatic increase in network connectivity (middle panels in Biii and C). In the case of complete silencing, the affected neurons disconnected from the network (right panels in Biii and C).

Divergent regulation of network connectivity under stimulation and deprivation via three structural plasticity rules.

(A) A subpopulation (S) consisting of 10% excitatory neurons (E) was subject to activity perturbation. All E-E, S-S, and E-S synapses were governed by the biphasic Gaussian rule. (Bi) Activity perturbation protocol. Three different folds of the original intensity (FOI) from the Poisson generator were used as examples to represent stimulation (110% FOI), mild deprivation (95% FOI), and silencing (0% FOI). (Bii) Temporal dynamics of neural activity of S, E, and inhibitory neurons (I) under each protocol. (Biii) Temporal evolution of the overall network connectivity (Γoverall) and subgroup connectivity (Γsubs) under each protocol. The synaptic connection probability from E to S is identical to that from S to E, as the same rules were applied to spines and boutons. Therefore, only S-E traces are shown unless otherwise stated. (C) Network connectivity matrices at the end of the three protocols. (D) Average incoming synapse numbers of S neurons under different FOIs. The empty green circles represent data from networks subjected to extreme stimulation or inhibition, where both neural activity and network connectivity were unstable. (E) Examples of two neurons receiving the same external inputs but with different calcium decay time constants (τCa). The upper panel shows membrane potential; the middle panel displays spike trains; the lower panel depicts the integrated calcium concentration over time. (F) Connectivity traces of the subnetwork under silencing across three different conditions. The dashed circle area is shown at higher magnification.

A systematic analysis of incoming synapse numbers confirmed the biphasic dependency of the biphasic Gaussian rule (dark yellow curve in Fig. 5D) and the monotonic behavior of the linear rule (green curve). The stable Gaussian rule, which has two stable setpoints (η = 0), exhibited an intermediate response: external stimulation and mild deprivation triggered a homeostatic reduction or increase in synapse numbers, but strong deprivation left the network silent and intact (light yellow curve). Of the three rules, the biphasic Gaussian rule best captured the homeostatic properties of the linear rule while also allowing for non-homeostatic, silencing-induced spine loss, consistent with our experimental findings and previous studies.55–59

Lesion-or denervation-induced plasticity is of particular interest in activity-dependent structural plas-ticity and has clinical relevance.60–62 The biphasic Gaussian rule provides a useful model for studying denervation-induced degeneration and subsequent regeneration. However, we observed a small, biologi-cally unrealistic increase in connectivity immediately following silencing, possibly due to a fast growth rate or residual effects of prior activity on intracellular calcium concentrations. In Fig. 5E, we showed that two neurons receiving the same external inputs displayed identical membrane potential dynamics and spike trains, but difference in the calcium time constant (τCa) caused a divergence in accumulated calcium concentration. Since synaptic element numbers depend on intracellular calcium concentration in our model, which reflects firing rate, a shorter time constant (τCa = 1 s, orange curve in 5F) carried less activity history, leading to a smoother reduction in synaptic connectivity. A similar effect was achieved by slowing the growth rate (10% of the original rate) for structural plasticity (light blue dashed curve, 5F). Our computer simulations showed that the timescales of calcium dynamics and structural plasticity interact. The decay time constant for calcium concentration is difficult to estimate due to the diversity of calcium signals, which vary in time and spatial compartmentalization. Additionally, structural plasticity is typically a slow process, occurring over minutes, hours, and days. Therefore, we used the slow growth rate and the original calcium time constant for subsequent studies. The faster growth rate employed in Fig. 6D-F was used solely to accelerate spine turnover, thereby reducing simulation time and allowing us to observe long-term changes more quickly.

Activity perturbation and recurrent connectivity shape the evolution of network connectivity.

(A) In the default network, the same growth rule was applied to both axonal boutons (A, light brown curve) and dendritic spines (D, pink curve), with ηA = ηD, as shown in the upper inset. Different η values could be used for axonal and dendritic elements, such as ηA > ηD or ηA < ηD. A silencing protocol was applied (fold of original intensity, FOI). (B) Neural activity of the stimulated subpopulation (S), excitatory neurons (E), and inhibitory neurons (I) under two conditions. (C) Network connectivity matrices at the conclusion of the silencing protocol under the two conditions. (D) Protocols used to investigate the effects of recurrent connectivity and external stimulation. The same growth rule was applied to both axonal and dendritic elements (ηA = ηD) here. (E, F) Time courses of neural activity and connectivity following silencing and external stimulation. In Protocol 1, the growth rate for the left panel (Prtcl. 1; left panel) was ten times faster than in the left panel (Prtcl. 1; right panel). In Protocol 2 (Prtcl. 2), the intensity of external stimulation was doubled as compared to the one used in Prtcl. 1.

Recurrent connectivity and activity perturbation shape spine number evolu-tion during synaptic recovery

While silencing-induced disconnection captured the characteristic spine loss observed during input de-privation, complete neural isolation is rarely observed in vivo. Instead, spine recovery often occurs following an initial loss of spines.55, 57, 59 Previous computer simulations using a Gaussian rule have also reported such “physiological” recovery, assuming distinct growth rules for axonal bouton and dendritic spines.43, 44 However, when we applied different η values for axonal boutons and dendritic spine growth within a topology-free network (without distance-dependent connectivity), synapse numbers failed to recover after silencing (Fig. 6A,B). Instead, we observed only asymmetric connectivity in the input and output synapses of the deprived subpopulation (Fig. 6C). A closer examination of the network revealed the critical role of recurrent inputs. In networks with distance-dependent connectivity, neurons near the boundary of the deprived region continued receiving active inputs from non-deprived neighbors, making them less deprived than neurons at the center, even though all external inputs were equally removed.

This observation led us to further investigate the role of recurrent inputs in synaptic recovery. In our topology-free network, we assessed the effects of recurrent inputs by adjusting both external inputs and internal connectivity. After applying the silencing protocol, we introduced external stimulation to the deprived subnetwork with different recurrent connectivities. In Fig. 6E,F (Protocol 1), the network in the left panel was simulated at a speed ten times faster than in the middle panel, meaning that the recurrent connectivity was much lower in the left panel when stimulation was applied. To simplify the analysis, we avoided complicating factors related to stimulation intensity – such as stimulation magnitude, du-ration, interval, and session numbers – that have been studied previously.19, 49, 63 Instead, we kept the stimulation constant and used two different intensities as a proof of concept. The results demonstrated that network dynamics are determined by the product of external inputs and internal connectivity. In the left panel of Fig. 6E, the deprived subnetwork remained silent despite stimulation, due to low recur-rent connectivity. In contrast, the same stimulation reactivated the deprived subnetwork in the middle panel, where connectivity was higher. Connectivity traces in Fig. 6F confirmed that residual recurrent connectivity amplified the external stimulation, leading to synapse regeneration and rewiring. However, if the external stimulation was too strong (protocol 2), the amplification effects of recurrent connectivity caused the system to shift from silencing-induced degeneration to over-excitation-induced homeostatic degeneration (Fig. 6E,F, right panels). These simulation results suggested that recurrent connectivity, in combination with external inputs, jointly determines neural activity and influences synapse numbers and network connectivity.

Synaptic scaling shapes intrinsic connectivity and modulates structural plas-ticity during activity perturbation

In line with this concept, homeostatic synaptic scaling could theoretically reshape recurrent connectivity and modulate neural activity by dynamically adjusting functional synaptic transmission. Specifically, homeostatic downscaling of excitatory synapses might reduce heightened neural activity, potentially masking changes in structural plasticity, particularly if it operates quickly and effectively. Conversely, homeostatic strengthening of excitatory synapses could increase neural activity. Under conditions of silencing, this strengthening may shift activity-dependent structural plasticity from non-homeostatic, silencing-induced synapse loss to homeostatic rewiring, similar to the outcomes observed with external stimulation or distance-dependent connectivity.

To test this hypothesis, we implemented a monotonic, activity-dependent synaptic scaling rule guided by the same intracellular calcium concentration as the structural plasticity rule. To ensure consistent baseline connectivity and avoid developmental issues, we allowed the network to grow using the biphasic Gaussian structural rule until it reached the equilibrium state. We then activated the calcium-based synaptic scaling at the onset of the activity perturbation protocol (Fig. 7A; Supplementary Fig. 4A). To disentangle the contributions of structural plasticity and synaptic scaling, we analyzed two types of connectivity: (1) Structural Connectivity (Γstruc.), determined solely by the number of synapses; and (2) Effective Connectivity (Γeffec.), calculated by multiplying the number of synapses by the weight of each synapse. Without synaptic scaling, synaptic weights remained uniform, resulting in identical values for structural and effective connectivity (Fig. 7C, left panels; Supplementary Fig. 4Bi). Intuitively, when neural activity was increased through stimulation, both structural plasticity and synaptic scaling operated redundantly in a homeostatic manner, leading to a less pronounced homeostatic disconnection in the presence of synaptic down-scaling.

Homeostatic synaptic scaling (HSS) shapes effective connectivity and interacts with homeostatic structural plasticity (HSP).

(A) Protocol of silencing – 0% FOI, applied to a subpopu-lation of excitatory neurons (S) – and synaptic scaling activation. Three different scaling strengths were applied: ρ = 0 represents no scaling, ρ = 0.01 represents weak scaling, and ρ = 0.02 represents strong scaling. (B-C) Time courses of network activity and connectivity. Γstruc. refers to synapse-number-based structural connectivity, while Γeffec. represents effective connectivity, which is the product of synapse numbers and synaptic weights. (Di-Dii) Structural and effective connectivity matrices of the entire net-work at t2. (Ei-Eii) Structural and effective connectivity matrices of the entire network at t3. (F) Raster plots for 100 selected neurons, including inhibitory (red, I), excitatory (blue, E), and silenced excitatory neurons (grey; S), from the network at the time points t1, t2, and t3.

In contrast, activity deprivation involved a complex interplay between synaptic scaling and structural plasticity. With weak scaling (Fig. 7B,C, ρ = 0.01), there was a gradual increase in effective connectivity to the deprived neurons (E-S), but this had minimal impact on firing rates and was insufficient to restore synapse numbers and network connectivity. Increasing the scaling strength (ρ = 0.02), as shown in the right panel of Fig. 7C, resulted in a more pronounced increase in effective connectivity in E-S synapses, which successfully reactivated the deprived neurons from their silent status, initiating synapse regeneration and rewiring. We also examined the structural and effective connectivity matrices at time points t2 (Fig. 7Di,Dii) and t3 (Fig. 7Ei,Eii). Notably, while the number of input synapses to the deprived subgroup decreased at t2, the remaining synapses that received inputs from the non-deprived neurons increased in synaptic weight. This reactivated the silent neurons (Fig. 7B, right panels), and by t3, the number of input synapses to the deprived subgroup began to recover. This observation confirms that monotonic synaptic scaling operates in opposition to the biphasic structural plasticity rule under conditions of severe activity deprivation, facilitating a transition from non-homeostatic synapse loss to homeostatic rewiring. Intriguingly, with sufficient scaling strength, synapse numbers and average neural activity were restored following silencing, and the temporal dynamics of the network were also altered after synapse regeneration. The activity pattern of the previously deprived population showed high synchronization after recovery (Fig. 7F), highlighting mechanisms that may be relevant to post-traumatic conditions such as injury-triggered epilepsy. Overall, our simulation results suggested that biphasic homeostatic structural plasticity and monotonic homeostatic synaptic scaling serve complementary and redundant roles in maintaining firing rate homeostasis, providing a biologically plausible explanation scenario through their interplay.

Spine size dynamics suggest an interaction between structural plasticity and synaptic scaling

The altered dynamics in the deprived neural population after recovery may result from dominant inputs from non-deprived excitatory neurons. To test this, we analyzed the firing rates and input synaptic weights of representative neurons from both the deprived (grey) and non-deprived (blue) excitatory populations (Fig. 8A). This analysis revealed a notable increase in synaptic weights from non-deprived to deprived neurons (Fig. 8A, lower panel), which was further corroborated by the temporal changes in synaptic weight distribution. Additionally, we assessed input synapse numbers for the representative neuron (solid line) and its peers (dashed line in Fig. 8C) to gain insights into structural plasticity. Consistent with experimental findings, the input synapse numbers of the deprived neurons decreased following silencing but gradually recovered. Interestingly, recovery patterns differed: the synapse numbers from reactivated neurons exceeded initial values (upper panel), whereas those from non-deprived neurons (middle panel) and the total synapse counts (lower panel) restored to slightly lower levels than before the perturbation. Our simulation results demonstrated that an appropriate combination of synaptic scaling and structural plasticity can replicate the process of spine loss and recovery following activity deprivation. Importantly, the model aligns with the commonly observed outcome of chronic activity silencing: fewer but larger-weight synapses, corresponding to fewer but larger spines.

Spine size analysis suggests interactions between functional and structural plasticity.

(A) Activity trajectories and average weights of incoming excitatory synapses to two sample excitatory neurons, N1 (deprived) and N2 (non-deprived). The grey dashed line in the upper panel indicates the setpoint activity level. Incoming synapses were grouped into two categories: those from the deprived subpopulation (S, grey) and those from non-deprived excitatory neurons (E, blue). The triangles in the middle panel mark the time points at which synaptic weight distributions were analyzed in panel B. (B) Synaptic weight distributions of neuron N1 at two time points, labeled as before and after. The numbers indicate the total synapse count for each type at the corresponding time point. (C) Normalized synapse number of the neuron N1 (solid line) and the averaged value across all deprived neurons (dashed line). (Di-Diii) Baseline spine size distributions for three groups. (N = 489 for the control group, N = 736 for the 200 nM-NBQX group, N = 675 for 50 µM-NBQX group) (E) Cumulative distribution function (cdf) of spine sizes before and after the three-day treatment. The inset shows the corresponding averages of normalized spine sizes. (F) Normalized changes in spine sizes, grouped by their initial spine sizes. The values on the x-axis represent the upper limits of each binning group. Data points above the dashed line indicate spine enlargement, while those below indicate spine shrinkage.

To verify if this principle holds in our NBQX experiment and to explore any dose dependency, we an-alyzed spine sizes before and after the chronic NBQX treatment in entorhinal-hippocampal slice cultures. Initially, spine sizes in all three groups showed a long-tail distribution, with a predominance of small spines (Fig. 8Di-Diii). After the three days of treatment, changes in spine sizes (Δ spine size) followed a normal distribution (Supplementary Fig. 3Bi-Biii). At the population level (Fig. 8CD), significant increases in spine sizes were observed following 200 nM (p < 0.001, Wilcoxon test; 99.9%CI = [9.09, 25.8], LLM), while 50 µM NBQX treatment lead to significant reductions (p < 0.001, Wilcoxon test; 99.9%CI = [9.09, 25.8], LLM), suggesting a non-monotonic, activity-dependent response. When normalizing the size changes by initial spine size, we observed a general increase across all three groups (Fig. 2D inset), likely driven by the enlargement of many small spines. Sorting the spine size changes by initial sizes (Fig. 2E) showed that in control segments (black curve), small spines generally grew while large spines shrank over the three-day period. Under partial inhibition with 200 nM NBQX (orange curve), the dynamics favored an overall increase in spine sizes, regardless of initial sizes. However, with complete inhibition using 50 µM NBQX (light blue curve), only a small number of large spines increased in size, counteracting the natural growth of small spines and causing medium-sized spines to shrink.

These spine data suggested a general homeostatic adjustment of spine sizes in response to partial and complete inhibition. Notably, under complete inhibition, the spine size changes were dependent on initial spine sizes, mirroring experimental findings and our simulation results, where complete activity blockade resulted in fewer but enlarged spines. However, this pattern does not fully align with synaptic weight changes described by the synaptic scaling rule. This discrepancy will be further explored in the discussion, where we integrate electrophysiological and imaging data to provide a comprehensive interpretation.

Hybrid combinations of synaptic scaling and structural plasticity resolve spine density inconsistencies

Having obtained both spine density and size data for individual segments, we explored whether a universal rule might link spine density and size changes across dendritic segments. We quantified the average spine size changes (Δ̅ spine size) for each segment and correlated these with baseline spine densities (Fig. 9A) and changes in spine density (Δ spine density, Fig. 9B). A significant positive correlation between spine size changes and baseline spine densities was observed in the 200 nM NBQX treatment group (Fig. 9A, r(24) = 0.65, p = 0.00064, Pearson’s correlation; Supplementary Fig. 2A), but this correlation was absent in the control and 50 µM NBQX groups. When correlating average spine size changes with alterations in spine density across all segments, weak inhibition (200 nM NBQX) typically resulted in increases in both spine density and size in most segments (see orange dots clustered in the upper right quadrant in Fig. 9B). This suggested that spine-dense segments may play a key role, with a synergistic interaction between structural and functional plasticity under weak inhibition. In contrast, under strong inhibition with 50 µM NBQX, the size-density data points were more widely scattered across the four quadrants (light blue dots in Fig. 9B), reflecting the diverse spine size modifications observed. Although our imaging data did not reveal a universal rule applicable to all segments, the results suggested a highly variable response to activity deprivation across individual dendritic segments. This led us to hypothesize that different combinations of synaptic scaling and homeostatic structural plasticity may operate at the dendritic segment level or even vary across different cell types (Fig. 9C). This diversity may explain the inconsistent findings observed in neurons from different regions or the contrasting effects seen between apical and basal dendrites within the same neuron.64

Systematic study of the interaction between synaptic scaling and structural plasticity in response to activity silencing.

(A) Correlation between each dendritic segment’s average change in spine size and its initial spine density. Marker size represents the net change in spine density over a three-day period. (B) Correlation between each dendritic segment’s average changes in spine size and its change in spine density. (C) Hypothesis: Different combinations of homeostatic synaptic scaling (HSS) and homeostatic structural plasticity (HSP) may form a spectrum, allowing different neuron types or dendritic segments within the same neuron to be governed by unique subsets of these rules. This spectrum could explain the high diversity observed in empirical studies of structural plasticity. (Sample neurons were reconstructed based on a CA1 pyramidal neuron (left two) and a dentate gyrus granule cell (right two), previously recorded in our lab.) (D) Simulation protocol. We systematically varied the growth rate of the HSP rule (ν) and the scaling strength of the HSS rule (ρ). (Ei-Eii) Example traces of structural connection probability and neural activity of the deprived subpopulation (S) under different parameter combinations. (Fi-Fii) Discrepancies in connectivity and firing rates when different growth rates and scaling strengths were combined. Discrepancies were calculated by estimating the area between the actual time course and the equilibrium connection probability (10%) or the target rate (ɛ) from the time of silencing until the end of the simulation. Results are averaged from 11 random trials and displayed in panels E and F.

To explore this hypothesis, we applied the same silencing protocol in simulation, varying the growth rates of structural plasticity (ν) alongside different synaptic scaling strengths (ρ) (Fig. 9D). Example trajectories of neural activity and structural connectivity in the deprived subpopulation under different conditions are shown in Fig. 9Ei,Eii. To evaluate the effect of these parameter combinations on network rewiring, we summed the differences between actual structural connectivity in the subnetwork and the equilibrium state value (10%) over the entire silencing period (Fig. 9Fi). Cold colors represent a failure in connectivity restoration, while warm colors indicate successful rewiring or over-reconnection. Obser-vationally, slower growth rates led to gradual spine loss, resulting in more residual synapses. In this case, even weak synaptic scaling was sufficient for synapse recovery. In contrast, faster growth rates, associated with fewer residual synapses, required stronger synaptic scaling for comparable restoration.

We also tracked discrepancies between actual neural rates and the target rate (7.9 Hz) for the deprived population over the same period (Fig. 9Fii). Parameter combinations that failed to restore network connectivity (Fig. 7Fi) showed negative deviations from the target rate (dark blue in Fig. 7F), indicating that the reactivation of silent neurons was unsuccessful. However, successful synapse regeneration (warm colors in Fig. 7Fi) did not always align with smooth firing rate restoration. Optimal firing rate restoration was observed with parameters leading to moderate synapse regeneration (light pinkish colours in panel Fii). Combinations of rapid structural growth and strong synaptic scaling could lead to excessively high neural activity (bright colors), mimicking epileptiform activity, or result in decreased transient firing rates (dark blue colors). In conclusion, our systematic study confirmed that calcium-based synaptic scaling and homeostatic structural plasticity interact across varying time scales and strengths. Their interplay helps reconcile the disparate experimental findings reported in previous studies.50

Discussion

The mammalian brain, a complex system with billions of neurons and non-neuron cells organized into layers, regions, and circuits,65 is highly robust such that flooding information does not cause catastrophic forgetting,66 nor will input deprivation destroy the network dynamics.15–17 Hebbian plasticity and home-ostatic synaptic scaling (HSS), both well-studied forms of functional plasticity, are central to maintaining this neural network’s functional robustness and dynamics. However, structural plasticity has been rel-atively undervalued due to inconclusive experimental evidence50 and limited computational models.42 Under activity deprivation, both homeostatic and non-homeostatic changes in spine numbers have been observed,50 posing a challenge to the promising concept of the homeostatic structural plasticity (HSP) model. This manuscript details the biphasic HSP model and its interaction with the monotonic HSS under activity deprivation. Our time-lapse imaging experiments confirmed the non-monotonic activity dependency of spine numbers, supporting the biphasic HSP rule. Simulations demonstrated that this rule reconciles homeostatic and non-homeostatic changes, complementing the HSS rule in adjusting synaptic weights and influencing synapse numbers. The redundancy and heterogeneity between HSP and HSS rules are essential for maintaining firing rate homeostasis and ensuring the stability and functionality of brain networks.

Our research significantly expands the understanding of structural plasticity by providing evidence of its differential, activity-dependent regulation. Although structural plasticity is often associated with functional cortical reorganization following injury and amputation or in cases of disease-related cognitive decline, its modeling has been less extensive compared to functional plasticity due to the challenges of capturing high-dimensional morphological changes. Among the existing mathematical implementations, some focused on dendritic tree development,67–69 while others emphasized synapse formation and rewiring as the key differentiator from functional plasticity models. However, these early models mostly adopted a synapse-specific perspective akin to Hebbian plasticity, emphasizing used-based growth and removal of individual synapses,70–74 sharing limitations in stabilizing network dynamics. In contrast, Butz and van Ooyen introduced a cell-autonomous HSP rule, rooted in van Ooyen and van Pelt’s work75 and aligned with concepts proposed by Dammasch.76 This rule employs a setpoint firing rate or calcium concentration to guide spine and bouton formation or elimination.43, 77 This spine-number-based rule is a structural analog to synaptic scaling in maintaining firing rate homeostasis. In the meantime, it retains the associative properties of correlation-based models19, 46, 47 and promotes self-organized criticality in developing neural networks.78 Therefore, in linear or Gaussian format, this rule has long been suggested as an alternative solution to integrate Hebbian and homeostatic plasticity, replicating various phenomena observed in development and disease-like conditions. Our work supports the biological plausibility of a biphasic HSP rule, substantiating it with experiments and computational evidence.

We used gradual activity blockade and time-lapse imaging to propose that spine numbers follow a biphasic relationship with neural activity, addressing inconsistencies in spine density findings across var-ious studies.50 Experimental paradigms, such as lesion methods and pharmacological treatments, often introduce confounding factors like neuroinflammatory cytokine production,57, 58, 79, 80 which influences plasticity.81–83 Non-traumatic deprivation methods, such as dark rearing84 or whisker trimming,85 result in non-uniform input deprivation across the dendritic tree, complicating interpretations of spine den-sity changes across different dendritic compartments. Precise pharmacological treatments, such as those targeting pre-synaptic mechanisms86 or using glutamate uncaging near selected spines,87 also have limita-tions due to the homogeneity of their effects. Even studies using drugs that uniformly target postsynaptic neurons report inconsistent results in spine density,88–93 which may have suggested the potential role of the strength of deprivation. Factors like drug type (non-competitive or competitive), experimental setting (in vivo or in vitro), and circuit integrity significantly influence the degree of inhibition experienced by target neurons. Our recent findings in CA1 pyramidal neurons revealed distinct cellular, structural, and molecular responses to synaptic input deprivation depending on the pathway involved, highlighting the pathway-dependent nature of these changes.94 Based on these observations, we hypothesized a bi-phasic activity-dependent rule to explain the wide range of spine density changes: partial activity suppression promotes spine growth, while complete inhibition results in spine loss. Indeed, our time-lapse imaging confirmed this hypothesis, showing increased spine density with partial inhibition (200 nM NBQX) and decreased density with complete inhibition (50 µM NBQX). However, the results also pointed to com-plex behaviors in spine size upon inhibition, suggesting an interaction between functional and structural plasticity, which we will explore further in the discussion.

The biphasic structural plasticity model, in the shape of a Gaussian growth curve, successfully rec-onciled the diverse activity-dependent changes of spine density observed in experimental settings in our simulations. With an unstable setpoint (η > 0), this model predicts variable outcomes in synapse numbers (corresponding to spine numbers)–ranging from increases to no changes or reductions–following activity perturbation. Notably, synapse loss induced by silencing falls outside the predictive scope of linear (mono-tonic) models but is captured by the biphasic rule. As a derived feature, this biphasic rule enables phase transition by modulating neural activity through external input or intrinsic connectivity adjustments. For example, timely and appropriately scaled external stimulation can reverse silencing-induced synapse loss, leading to homeostatic rewiring. This insight could inform non-invasive brain stimulation techniques aimed at preventing the negative effects of stroke-induced neural lesions.95 Additionally, the homeostatic response of synaptic scaling promotes recovery of synaptic connections, consistent with empirical obser-vations of synapse numbers and weights. This natural rewiring process has also been observed in studies by Butz and van Ooyen using the same Gaussian HSP rule,41, 43, 44 where they applied distance-dependent connectivity and emphasized the importance of using distinct growth rules for axonal boutons and den-dritic spines. Our successful replication of this rewiring in a topology-free homogeneous network supports their findings, highlighting the importance of intrinsic connectivity in regulating synapse loss and network rewiring while freeing the network from architectural constraints. However, our co-simulation of func-tional and structural plasticity rules revealed a critical yet synergistic interplay between these different forms of plasticity. This interaction, often overlooked, is essential for interpreting experimental data or modeling biological systems. Moreover, inhibitory plasticity may contribute to the physiological recovery of homeostatic structural plasticity, as suggested by numeric studies,96 though this was beyond the scope of our study and warrants future exploration.

In this context, we examined how the tandem behavior of homeostatic structural plasticity and func-tional synaptic scaling shapes our interpretations. Our experiments revealed an increase in both spine size and density following three days of 200 nM NBQX treatment. This enlargement of spine size likely increased the detectability of small spines, suggesting that the observed increase in spine density re-sults from a combination of spinogenesis, spine stability, and the potentiation of smaller spines. Similar confounding factors affect the interpretation of synaptic scaling results. For instance, Turrigiano et al.26 observed changes in miniature EPSC amplitudes but not the frequencies after chronic inhibition and excitation, inferring synaptic weight scaling. However, other studies have reported changes in both amplitudes and frequency. The transformation of silent synapses into active ones97 can increase the frequencies but lower the average EPSC amplitudes due to smaller synaptic weights. These represent synapse-number-based structural plasticity, which may not always be detectable with conventional imag-ing techniques, depending on factors like laser intensity or exposure time. Additionally, under complete inhibition, we observed distinct size changes in small and large spines: smaller spines tend to shrink and disappear, while larger spines enlarge, leading to increased sEPSC amplitudes. This could result in different or even contradictory results between electrophysiological recording and imaging experiments. Therefore, spine numbers and synaptic weights may represent two facets of a broader concept–“effective transmission”–and focusing on only one aspect risks missing the complete picture.

Our systematic study successfully replicated the empirical observation that chronic activity inhibition often leads to spine loss, with remaining spines exhibiting enlarged heads.90, 91 This phenomenon can be attributed to the shared calcium signaling mechanism underlying both HSP and HSS rules. While the HSP rule consistently incorporated activity-based calcium concentration as a setpoint, most HSS models focus on firing rate setpoint,98–101 synaptic weight normalization,102 or presynaptic regulation.103, 104 Notably, one model proposed an integral feedback control algorithm,51 using a variable with an exponential kernel to track neural spiking activity and adjust synaptic weights. At the time, the role of calcium concentration as the key variable tracking neural activity was not yet known.105 Leveraging this model, we reinterpreted the internal status using a calcium-based variant to explore the interaction between the HSS and the HSP rules. As a result, our study also served as a thought experiment (Gedankenexperiment) to understand the interplay between functional and structural plasticity. If both are redundantly monotonic, the rule with the faster timescale would dominate. However, when a biphasic HSP rule interacts with a monotonic HSS rule, bifurcation occurs under strong inactivity, as confirmed by our simulation, resulting in neurons with reduced spine density and enlarged spine heads. From an energy-efficient perspective, synapse formation and spine growth are constrained by protein availability and ATP consumption. Thus, it is more efficient for neurons to potentiate existing synapses rather than forming new ones. The combined effects of synaptic scaling and structural plasticity in maintaining network activity are supported by recent theoretical work at the dendritic trafficking level.106 Our study highlights the interdependency between these two forms of plasticity, which may explain the divergent results obtained across different brain regions, neuron types, dendritic segments, and experimental settings.

While our study addressed both synaptic scaling and homeostatic structural plasticity, it is important to note that the use of NBQX is not typical in experiments focused on synaptic scaling. NBQX, a competitive antagonist of AMPA receptors, allows for precise modulation of network activity through dose control,107 offering greater flexibility compared to tetrodotoxin (TTX), which blocks action potentials by inhibiting sodium channels. This makes NBQX an effective tool for modulating network activity, as demonstrated by our electrophysiological recordings and supported by other in vivo and in vitro studies.107, 108 However, the specific affinity of NBQX for GluA2-containing AMPA receptors, which play a key role in both TTX-induced synaptic scaling109 and glutamate-induced spine protrusion in the presence of TTX,110 warrants careful consideration. This selective binding raises questions about the broader implications and mechanisms of NBQX action, particularly in experiments where TTX is absent. While no definitive evidence suggests that NBQX alone impairs the synthesis or insertion of AMPA receptors, further research is needed to fully understand its effects on synaptic scaling and structural plasticity.

In summary, the use of calcium signal in both HSP and HSS rules resembles an integral feedback controller (Supplementary Fig. 5), a widely used engineering approach to achieve robust performance despite external perturbations. In both rules, calcium concentration serves as the control signal to drive connectivity kernel modification via negative feedback, which integrates neural firing rates over time and returns a filtered signal responding to its activity. However, the HSP and HSS rules differ in their detailed mechanism. The HSP rule more closely aligns with a Proportional-Integral-Differential (PID) controller, while the HSS rule operates more like a Proportional-Integral (PI) controller. Besides sharing the integral (I) component, the HSP rule’s proportional (P) component is linked to the growth rate of synaptic elements,45 which modifies the number of synaptic elements created or deleted per time unit based on a magnitude relative to the difference between the expected and actual calcium concentration. The scaling factor (ρ) in the HSS rule regulates the proportional effects of synaptic scaling. Uniquely, the HSP rule includes a derivative (D) component, determined by the steepness of the growth curve, which is determined by the steepness of the growth curve and dictates how quickly or slowly the system reaches the desired neural activity. This derivative component is critical for ensuring a smooth and robust response to stimuli, avoiding biologically unrealistic overshooting and abrupt oscillations. This is particularly relevant in the Gaussian case, where the distance to the target firing rate accelerates differently depending on the position on the calcium concentration axis. Thus, although the exact shape of the growth rule for homeostatic structural plasticity is still uncertain, we propose that a biphasic curve with a changing slope provides a redundant and heterogeneous structural backup for synaptic scaling, helping maintain robust firing rate homeostasis. This perspective also helps anticipate interactions with other plasticity rules where sub-or supra-threshold calcium dynamics are involved.111, 112

Materials and methods

Ethics statement

We used mouse pups postnatal at 3 to 5 days (P3-P5) from C57BL/6J (wild-type) and Thy1-eGFP mouse lines to prepare entorhinal-hippocampal tissue cultures in the current study. All animals were kept under a 12 h −12 h light-dark cycle with food and water provided ad-libitum. One male and one or two female(s) were kept within the same cage for breeding. All animal experiments were approved by the appropriate animal welfare committee and the animal welfare officer of Albert-Ludwigs-University Freiburg, Faculty of Medicine under X-21/01B and X-18/02C. All effort was made to reduce the pain or distress of animals.

Preparation of tissue cultures

Entorhinal-hippocampal tissue cultures were prepared as published before.113, 114 All tissue cultures were cultivated for at least 18 days inside the incubator with a humidified atmosphere (5% CO2 at 35 ◦C to reach an equilibrium status. The incubation medium consists of 50% (v/v) 1× minimum essential media (#21575 − 022, Thermo Fisher, USA), 25% (v/v) 1× basal medium eagle (#41010 − 026, Thermo Fisher, USA), 25% (v/v) heat-inactivated normal horse serum, 25 mM 1 M HEPES buffer solution (#15630 − 056, Gibco), 0.15% (w/v) sodium bicarbonate (#25080 − 060, Gibco), 0.65% (w/v) glucose (#RNBK3082, Sigma), 0.1 mg/ml streptomycin, 100 U/ml penicillin, and 2 mM glutamax (#35050 − 061, Gibco). The incubation medium was renewed 3 times per week. If not stated otherwise, the incubation medium applied to the tissue cultures was always pre-warmed to 35 ◦C and adjusted around pH = 7.38.

Experimental design

The main objective of the experiments in the current study was to probe whether there is a linear or non-linear dose-dependent regulation of dendritic spine density such that we can implement a spine-number-based structural plasticity rule close to biological reality for further systematic studies. Compet-itive AMPA receptor antagonist NBQX (2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline) at different concentrations was bath applied to wild-type cultures while recording synaptic transmission at CA1 pyramidal neurons. Two magnitudes of inhibition (“partial” and “complete”) were thus determined by the reduction of amplitudes and frequencies of spontaneous excitatory postsynaptic currents (sEPSCs). Then the same concentration ladders were used to treat Thy1-eGFP cultures for three days, where we tracked individual dendritic segments of CA1 pyramidal neurons before and after the three-day treat-ment to investigate whether two different magnitudes of AMPA-receptor inhibition resulted in divergent alterations of spine densities and spine sizes.

Whole-cell patch-clamp recordings

To probe the effects of NBQX administration at different concentrations on synaptic transmission, whole-cell patch-clamp recordings were conducted in CA1 pyramidal neurons. Recordings were performed at 35◦C. The bath solution contained (in mM) 126 NaCl, 2.5 KCl, 26 NaHCO3, 1.25 NaH2PO4, 2 CaCl2, 2 MgCl2, and 10 glucose (aCSF) and was continuously oxygenated with carbogen (5% CO2/95% O2). Glass patch pipettes had a tip resistance of 4 − 6 MΩ, filled with the internal solution which contained (in mM) 126 K-gluconate, 10 HEPES, 4 KCl, 4 ATP-Mg, 0.3 GTP-Na2, 10 PO-Creatine, 0.3% (w/v) biocytin. The internal solution was adjusted to pH = 7.25 with KOH and reached 290 mOsm with sucrose). We patched 6 neurons per culture to record the spontaneous excitatory postsynaptic currents (sEPSCs) of CA1 pyramidal neurons in the voltage-clamp mode at a holding potential of −70 mV. Series resistance was monitored before and after each recording and the neuron data was excluded if the series resistance went above 30 MΩ. Each neuron was recorded for 2 min.

NBQX treatment

NBQX (2,3-dioxo-6-nitro-7-sulfamoyl-benzo[f]quinoxaline) is a competitive antagonist of AMPA recep-tors.115 We chose two concentrations 200 nM and 50 µM and delivered by bath treatment to achieve a partial or complete inhibition of AMPA receptor currents. Wild-type cultures were either recorded in normal ACSF or ACSF that contained two different NBQX concentrations. In time-lapse imaging experiments, Thy1-eGFP cultures were treated with 200 nM and 50 µM by adding NBQX (Cat. No. 1044, Tocris Bioscience, Germany) in the incubation medium for three days. Whenever we changed the new medium, fresh NBQX was administrated accordingly.

Tissue fixation and immunohistochemical staining

Recorded cultures were fixed and stained for post hoc inspection. Cultures were fixed by immersing into 4% (w/v) paraformaldehyde (PFA) in 1× phosphate-buffered saline (PBS, 0.1 M, pH = 7.38) for 1 h and transferred into 1× PBS for storage at 4 ◦C after being washed in 1× PBS. Before staining, all fixed cultures were again washed three times with 1× PBS (3 × 10 min) to remove residual PFA. We incubated the cultures with Streptavidin 488 (1 : 1000, #S32354, Invitrogen, Thermo Fisher, USA) in 1× PBS with 10% (v/v) in normal goat serum and 0.05% (v/v) Triton X-100 at 4 ◦C overnight. In the next morning, cultures were rinsed with 1× PBS (3 × 10 min) and incubated with DAPI (1 : 2000) in 1× PBS for 20 min. After another 4 washes with 1× PBS (4 × 10 min), we mounted the cultures on glass slides with DAKO anti-fading mounting medium (#S302380 − 2, Agilent) for confocal microscope imaging.

Time-lapse imaging

To inspect whether neural spine densities and spine sizes were altered by the three-day administration of NBQX, we employed time-lapse imaging to follow the same apical dendritic segments of CA1 pyramidal neurons before and after treatment. Live cell imaging was performed at a Zeiss LSM800 microscope with 10× water-immersion (W N-Achroplan 10×/0.3 M27; 420947-9900-000, Carl Zeiss) and 63× water-immersion objectives (W Plan-Apochromat 63×/1,0 M27; 421480-9900-000, Carl Zeiss). Thy1-eGFP tissue cultures where clear CA1 pyramidal neurons could be identified were used in this experiment. Dendritic segments from the radiatum layer were imaged. We imaged individual dendritic segments prior to the NBQX treatment and again after the three-day treatment. During the imaging session, the membrane insert with 4 cultures was placed into a 35 mm petri dish filled with 5 ml incubation medium on a platform which was constantly maintained at 35 ◦C. We used the same pre-warmed and pH-adjusted incubation medium for imaging procedures.

Experimental data quantification

Spontaneous excitatory postsynaptic currents (sEPSCs) were analyzed using the automated event detec-tion tool from the pClamp11 software package as previously described.116

z-stacked fluorescent images of Thy1-eGFP cultures were projected to create a 2D representation of individual dendritic segments. ImageJ plugin Spine Density Counter117 was used to count spine numbers and measure segment length, which estimates spine density. For the same dendritic segments imaged at different time points, special attention was paid to ensure the same starting and ending points were used. Post hoc visual inspection was applied to ensure the spine detection results were not strongly biased. Both raw spine density and normalized spine density by baseline were used in the analysis.

The same z-projected fluorescent images were used to track individual spines for spine size analysis. To eliminate the bias from drawing and automatic reconstruction, we drew circles manually around the spine to cut it from the dendrite; the spine size was estimated by measuring the signal intensity with an arbitrary unit of the drawn circle. The drawing and measurements were performed with FIJI ImageJ. Both the raw values and normalized values by baseline spine size were used in the analysis. Statistical methods were specified in the individual results section.

In the following analysis, we normalized the alteration of spine sizes by their baseline size before treatment and binned the data by their initial sizes. We first fetched the spine sizes of individual spines pre-and post-treatment to calculate the alteration (Δspine size). Then, we normalized the alteration amount by the corresponding baseline spine size. We binned the spines by their baseline initial sizes in a third step. By sorting the spine ID in each binned group, we grouped the spine size alterations (Δspine size) by their initial spine size. This analysis shows us how spines with different initial sizes update their size over a certain time course with and without NBQX treatment.

Statistical analysis

Dunn’s multiple comparison test was applied for statistical analysis regarding the sEPSC events among the three groups. For spine density analysis, the Wilcoxon test was applied to compare the values of each segment before and after the three-day treatment. If not otherwise stated, “ns” means no significant, “*” means p < 0.05, “**” means p < 0.01, “***” means p < 0.001. For spine size analysis, we first applied the Wilcoxon test for individual spines. To account for data clustering within each segment, linear mixed model (LMM) with spine_size timing was applied to compare the values before and after treatment with the segment ID as group factor as used in.48 The significance of LLM results was judged by whether the confidence interval (CI) crossed zero.

Neuron model and network model

We used the same spiking neuron model and network architecture as described before in46 and in.19, 48 Current-based leaky integrated-and-fire point neuron was used for both excitatory and inhibitory neurons. We build an inhibition-dominated network with 10 000 excitatory neurons and 2 500 inhibitory neurons.118 To simplify the scenario, we only grow the connections within the excitatory population (E-E) with the activity-based structural plasticity rule (see the Structural plasticity rule section below). Each inhibitory neuron was beforehand hard-wired randomly to receive synapses from 10% of the excitatory and inhibitory population. All details and parameters concerning neural and network models can be found in the Supplementary Materials. We performed the network simulations with NEST simulator 2.20.2 and NEST 3.0119 and MPI-based parallel computation. All the model parameters and protocols can be found in Supplementary Tables 1-7.

Structural plasticity rule

We enabled the growth, retraction, and rewiring of synapses among excitatory neurons with the help of structural plasticity rules. By definition, each excitatory neuron has multiple dendritic spines and axonal boutons, which are called synaptic elements. Synapses were formed by randomly matching free compatible synaptic elements. The growth and retraction of synaptic elements, or in other words, the number of synaptic elements, is governed by a growth rule. Three structural plasticity rules were explored in the current study: (i) linear growth rule; (ii) Gaussian growth rule with a zero setpoint and a non-zero setpoint; (iii) Gaussian growth rule with two non-zero setpoints. All three rules are determined by a function of calcium concentration that reflects neural activity, where C(t) is the time evolution of calcium concentration. Calcium concentration decays with a time constant τCa and increases with calcium influx (βCa) upon the emission of an action potential S(t) of the postsynaptic neuron. This operation is performed internally by the NEST simulator and works as a low-pass filtered signal of the spiking activity of the neuron. The growth of synaptic elements is regulated differently depending on the calcium concentration and the shape of the growth rule.

Linear growth rule

The linear rule was first introduced in45 and systematically studied in an inhibitory-dominant neural net-work.19, 46–48 The number of synaptic elements (z(t)) is linearly dependent on the calcium concentration, where ν is the growth rate, and ɛ is the target level of calcium concentration. Since calcium concen-tration reflects the neural activity loyally, the target level also suggests a setpoint of firing rate in the context of a certain neural network. As discussed before, when neurons fire below their target rate, they grow new synaptic elements and form new synapses. On the other hand, they break existing synapses and retract synaptic elements when they fire above the target rate (setpoint).

Gaussian growth rule

The Gaussian rule has a more complex dependency on the calcium concentration when neural activity is too low, as shown in Equation 3. This rule was suggested and explored initially in.43 where and . In this rule, η and ɛ are two setpoints: ɛ is the stable setpoint as used in the linear rule. When the neuron fires above ɛ, it retracts synaptic elements as in the linear rule. η is another setpoint introduced specifically for the Gaussian rule, determining the regulation manner when the neuron activity drops below ɛ. When the neuron is firing below ɛ but above η, the number of synaptic elements present will undergo homeostatic outgrowth, but when the neuron is firing below η, neurons will break synapses and retract elements. In the case where η = 0, neurons cease to change synaptic elements when their firing rate drops to zero.

Homeostatic synaptic scaling

In order to achieve homeostatic synaptic scaling, we make use of a new synaptic model in NEST called scaling_synapse. In this synapse model, the weight of the synapse is regulated by the difference between a homeostatic setpoint and the calcium trace of the postsynaptic neuron, where ρ is the scaling factor, and ɛ is the same target value as used in the homeostatic structural plasticity rule.

Activity perturbation

To examine different activity-induced scenarios on neural network connectivity, we performed systematic activity manipulation to a subnetwork of excitatory neurons (Nsub = 1 000), by changing its Poissonian input from 0% to 200% fold of the original intensity (FOI). All manipulations were performed at 6 000 s when the network has grown to the equilibrium state respectively with three structural plasticity rules. For the Gaussian rule with two non-zero setpoints, we applied damping current injection to the soma to facilitate its growth within the first 4 000 s and the network was simulated for another 2 000 s without any facilitating current.

Quantifying firing rate, network connectivity, and synapse number

Firing rate

Firing rate was calculated by the average spike count over a recording period for individual neurons. We have long interval (5 s) and short interval (1 s). Short intervals were only used within the short time window after activity perturbation to reveal its transient dynamics; otherwise, long intervals were used.

Network connectivity

Two types of connectivity were used in the present study. We used a N × N connectivity matrix (Aij) to represent the recurrent excitatory connections of our network, where columns and rows correspond to pre-and postsynaptic neurons. For structural connectivity, the entry Aij of the matrix represents the total number of synaptic connections from neuron j to neuron i. For effective connectivity, we integrated the synapse number with individual weights for each pair of neurons by summating the total weights. So the entry of the connectivity matrix is the equivalent number of unit synapses, by dividing the sum with a uniform weight 0.1 mV. To average the mean connectivity of the whole network or a subnetwork at any given time t, corresponding columns and rows of the connectivity matrix were selected and averaged by . Synapse numbers were calculated by the sum of the entry in the structural connectivity matrix.

Synapse number

Input and output synapse numbers, also called indegree and outdegree in another context, were calculated by summating the input and output synapse numbers of individual excitatory neurons based on the structural connectivity matrix.

Quantifying the discrepancies in firing rate and connectivity from the target values

To apply a systematic comparison among different combinations of the synaptic scaling strengths and structural plasticity growth rates, we summated the discrepancies in firing rate and connectivity from the target values for the subpopulation over time as an index of activity and connectivity recovery. All the discrepancies were calculated by estimating the area between the actual time course and the equilibrium connection probability (10%) or the target rate (ɛ) from the time of silencing until the end of the simulation. The method was explained in detail in Supplementary Fig. 6. To achieve the heat maps in Fig. 9, we averaged 11 random trials for each parameter combination.